The Bacterial Quorum-Sensing Signal 2-Aminoacetophenone Rewires Immune Cell Bioenergetics through the PGC-1α/ERRα Axis to Mediate Tolerance to Infection

Version 1

Abstract

Abstract

How bacterial pathogens exploit host metabolism to promote immune tolerance and persist in infected hosts remains elusive. To achieve this, we show that Pseudomonas aeruginosa (PA), a recalcitrant pathogen, utilizes the quorum sensing (QS) signal 2-aminoacetophenone (2-AA). Here, we unveil how 2-AA-driven immune tolerization causes distinct metabolic perturbations in macrophages’ mitochondrial respiration and bioenergetics. We present evidence indicating that these effects stem from a decrease in pyruvate transport into mitochondria. This reduction is attributed to decreased expression of the mitochondrial pyruvate carrier (MPC1), which is mediated by diminished expression and nuclear presence of its transcriptional regulator, estrogen-related nuclear receptor alpha (ERRα). Consequently, ERRα exhibits weakened binding to the MPC1 promoter. This outcome arises from the impaired interaction between ERRα and the peroxisome proliferator-activated receptor gamma coactivator 1-alpha (PGC-1α). Ultimately, this cascade results in diminished pyruvate influx into mitochondria and consequently reduced ATP production in tolerized macrophages. Exogenously added ATP in infected macrophages restores the transcript levels of MPC1 and ERR-α and enhances cytokine production and intracellular bacterial clearance. Consistent with the in vitro findings, murine infection studies corroborate the 2-AA-mediated long-lasting decrease in ATP and acetyl-CoA and its association with PA persistence, further supporting this QS signaling molecule as the culprit of the host bioenergetic alterations and PA persistence. These findings unveil 2-AA as a modulator of cellular immunometabolism and reveal an unprecedent mechanism of host tolerance to infection involving the PGC-1α/ERRα axis in its influence on MPC1/OXPHOS-dependent energy production and PA clearance. These paradigmatic findings paving the way for developing treatments to bolster resilience to pathogen-induced damage. Given that QS is a common characteristic of prokaryotes, it is likely that 2-AA-like molecules with similar functions may be present in other pathogens.

Introduction

Host tolerance is a fundamental mechanism of innate immunity. Studies have shown that innate immune cells following infection or exposure to microbial products may enter a state of immune tolerance characterized by reduced responsiveness towards microbial re-exposure a few hours later [1–3]. Epigenetic and signaling-based mechanisms have been proposed to be involved in host immune tolerance [1–3]. In recent years, several reports have highlighted the complex interplay between metabolic reprogramming and immunity [4, 5]. It was suggested that specific metabolic programs are activated in monocytes and macrophages upon exposure to infection and microbial products [6, 7]. However, in host tolerance, the molecular mechanisms underlying immunometabolic reprogramming mediated by bacterial pathogens remain poorly understood.

Metabolic pathways are essential for generating energy for various cellular functions, including those performed by immune cells [8, 9]. Macrophages use metabolic pathways to generate energy and metabolites to adapt to changing environments and stimuli, thereby enabling them to cope with the needs of a fluctuating immune response [7, 10]. Thus, properly functioning metabolic pathways are vital for immune cells to counteract pathogens. For instance, the crucial energy-carrying molecule adenosine triphosphate (ATP) is required to phagocytose pathogen-derived molecules efficiently.

Pseudomonas aeruginosa (PA), a recalcitrant ESKAPE (Enterococcus faecium, Staphylococcus aureus, Klebsiella pneumoniae, Acinetobacter baumannii, PA, and Enterobacter sp.) pathogen that causes acute and persistent infections, secretes virulence-associated low molecular weight signaling molecules, several of which are regulated by quorum sensing (QS) [11–14] and able to modulate host immune responses [15–18]. QS is a cell density-dependent signaling system bacteria use to synchronize their activities [11–14]. MvfR (a.k.a. PqsR), a critical QS transcription factor of PA, regulates the synthesis of many small molecules, including signaling molecules such as 2-aminoacetophenone (2-AA) [11, 19–21]. In vivo studies have demonstrated that 2-AA enables PA to persist in infected murine tissues by promoting innate immune tolerance through histone deacetylase 1 (HDAC1)-mediated epigenetic reprogramming [17, 18]. 2-AA tolerization reprograms the host inflammatory signaling cascade by maintaining chromatin in a “silent” state through increased HDAC1 expression and activity and decreased histone acetyltransferase (HAT) activity [18]. These changes also impact the protein-protein interaction between HAT and cyclic AMP response element-binding protein/HDAC1 and p50/p65 nuclear factor (NF)-κB subunits [18, 22]. The 2-AA-mediated tolerization permits PA to persist in infected tissues [17, 18] underscoring the difference from lipopolysaccharide (LPS)-mediated tolerization, which instead leads to bacterial clearance and involves different HDACs [23].

Mitochondria, the “powerhouse” of the cells, are crucial for the regulation, differentiation, and survival of macrophages and other immune cells [8]. Our group’s previous in vivo and in vitro studies indicated that 2-AA affects metabolic functions in skeletal muscle, which contains a high concentration of mitochondria [24–26]. Injection of 2-AA in murine skeletal muscle dampens the expression of genes associated with OXPHOS and the master regulator of mitochondrial biogenesis peroxisome proliferator-activated receptor-γ coactivator-1 beta (PGC-1β) [24]. Another PA QS molecule, 3-oxo-C12-HSL, attenuates the expression of PGC-1α [27]. PGC-1α and PGC-1β belong to the PPARγ family of inducible transcriptional coactivators and are known regulators of mitochondrial metabolism [28]. The coactivator PGC-1 proteins are involved in various cellular energy metabolic processes [29, 30], including but not limited to mitochondrial metabolism [31]. Recent studies have also emphasized the role of estrogen-related nuclear receptor α (ERRα) in coordinating metabolic capacity with energy demand in health and disease [32]. PGC-1α activates ERRα transcriptional activity through protein-protein interaction, which enhances the expression of ERRα [33, 34] and other mitochondrial genes involved in lipid metabolism and OXPHOS [35]. PGC-1α/ERRα axis has been extensively studied in cancer and established as a central regulatory node of energy metabolism that induces the global expression of genes involved in mitochondrial biogenesis and functions [36–38]. ERRα has been shown to occupy the promoter regions of many genes participating in the TCA cycle and OXPHOS [39, 40], including the mitochondrial pyruvate carrier (MPC1). In human renal carcinoma cells, PGC-1α/ERRα interaction results in efficient activation of MPC1 expression and transport of pyruvate into mitochondrion for efficient OXPHOS and energy production [41].

Given that our previous studies pointed to the 2-AA effect on mitochondrial functions [24–26] and that mitochondria are crucial for the regulation, differentiation, and survival of macrophages and other immune cells [8] we investigated how 2-AA may mediate cellular metabolic reprogramming in tolerized immune cells by examining the link between the PGC-1α/ERRα axis and 2-AA-mediated immune tolerance. Our findings uncovered the crucial action of 2-AA on energy homeostasis and metabolism in tolerized immune cells through perturbances of the ERRα/PGC-1α interaction, MPC1-mediated pyruvate transport and the production of the key energy metabolism molecules, ATP and acetyl-CoA and their association with the persistence of PA in mammalian tissues.

Results

Tolerization by 2-AA impacts the generation of crucial energy metabolites in macrophages

Following infection or exposure to microbial products, innate immune cells may enter a state of tolerance several hours later, as depicted by their diminished responsiveness or unresponsiveness to microbial re-exposure [1–3, 42]. Given that 2-AA tolerized cells are non-responsive to 2-AA re-exposure, we investigated the impact of 2-AA on energy homeostasis and metabolism in non-tolerized and tolerized BMDM cells (Fig. 1). We quantified the levels of ATP and acetyl-CoA, which is a fuel for ATP production, in a series of in vitro experiments in which murine BMDM cells were exposed to 2-AA (1st 2-AA exposure) or re-exposed (2nd 2-AA exposure) to respectively determine their activation, and tolerance as well as memory to this QS molecule (Fig. 1A). Initially, following the 1st 2-AA exposure (black bars), a significant increase in the concentrations of intracellular ATP (Fig. 1B) and acetyl-CoA (Fig. 1C) was observed in BMDM cells exposed to 2-AA for 1 and 6 h compared to their corresponding naïve control cells (grey bars). However, cells exposed to 2-AA for 48 h (black bar) appeared to have entered a state of tolerance as ATP and acetyl-CoA levels were decreased compared to 2-AA exposures for a shorter time (Figs. 1B and 1C). The tolerance of these cells as well as their memory to 2-AA was determined by their responsiveness to a 2nd 2-AA exposure (Fig. 1A). To achieve this, these cells were washed to remove 2-AA and allowed to rest for 24 or 106 h in the absence of 2-AA before receiving the 2nd 2-AA exposure for 1 or 6 h (Fig. 1A). Figures 1B and 1C (red bars) show the unresponsiveness of BMDM cells to 2-AA re-exposure as evidenced by the similar levels of ATP and acetyl-CoA compared to cells exposed to 2-AA for 48 h (1st exposure black bars). The sustained unresponsiveness of the cells to re-exposure indicates that the cells exhibited innate immunological memory, which could not be reverted by 2-AA re-exposure. Similar responsiveness patterns of ATP and acetyl-CoA as with BMDM cells were observed in murine macrophage RAW 264.7 cells (Figs. S1A and S1B) or human monocyte THP-1 cells (Figs. S1C and S1D) following exposure or re-exposure to 2-AA. These findings further support our observations with BMDM cells and confirm that either cell type can be used to study responses to 2-AA. These results reinforce the notion that 2-AA impacts crucial energy metabolites in macrophages.

2-AA tolerization decreases crucial metabolites of cellular energy and affects mitochondrial respiration in mouse BMDM.

(A) Schematic representation showing experimental design: naïve cells were exposed to 2-AA for 1, 6, or 48 h (black). Cells exposed to 2-AA (400 µM) for 48 h were washed, rested for 24 or 106 h, and re-exposed (200 µM) for 1 or 6 h (red), respectively. The same color code: black: cells after 1st exposure; red: cells after 2ndexposure, was kept throughout the manuscript with corresponding controls in grey and pink, respectively. The levels of (B) ATP and (C) acetyl-CoA in BMDM cells after 1st and 2nd 2-AA exposure. (D) Real-time oxygen consumption rate (OCR) traces were recorded using a Seahorse XF analyzer and normalized to protein content. Cells were exposed to 2-AA for 1 or 24 h (black), washed and rested for 24 h and re-exposed for 1 h (red). Mitochondrial respiratory parameters, (E) basal respiration, and (F) maximal respiration. Data are presented as mean ± SD, n ≥ 4, *p < 0.05, **p < 0.01, ***p < 0.001, and ns indicates no significant difference. One-way ANOVA followed by Tukey’s post hoc test was applied.

Moreover, we sought to determine whether 2-AA-tolerized BMDM cells would respond to heterologous stimulus. To test this, we re-exposed the cells to lipopolysaccharide (LPS), an outer membrane bacterial component that strongly induces macrophages. LPS re-exposure of the 2-AA tolerized BMDM cells did not significantly (p = 0.9) increase ATP levels (Fig. S2), indicating that 2-AA cross-tolerized the cells to this heterologous bacterial immunostimulant. On the other hand, as expected, we observed increased ATP levels following LPS stimulation of the non-tolerized BMDM cells (Fig. S2). The broad spectrum of 2-AA cross-tolerization strongly suggests an interplay of epigenetic mechanism rather than ligand-receptor mediated signaling-based mechanism in bringing out the observed 2-AA-mediated tolerance.

Tolerization by 2-AA leads to a quiescent state by reducing cell bioenergetics and compromising OXPHOS

To further investigate the 2-AA’s impact on the bioenergetics of macrophages, we assessed by Seahorse assay the oxygen consumption rate (OCR) as an index of OXPHOS in 2-AA exposed (for 1 and 24 h) and re-exposed (2nd exposure for 1 h) BMDM cells (Fig. 1D). Values of basal (Fig. 1E) and maximal (Fig. 1F) mitochondrial respiration and spare respiratory capacity (Fig. S3) interpreted from these measurements revealed that 1 h 2-AA exposure of naïve cells significantly increased OCR compared to naïve control cells. At 24 h post-2-AA exposure, cells had a significantly reduced basal OCR level than the control BMDM cells, indicating decreased OXPHOS and an overall quiescent phenotype of tolerized macrophages. Re-exposure of tolerized macrophages with 2-AA also did not augment maximal and basal OCR (Figs. 1D-1F), indicating the unresponsiveness of tolerized BMDM cells.

These OCR data are consistent with previous direct measurements of ATP (Fig. 1B) and acetyl-CoA (Fig. 1C) and support the notion that 2-AA tolerization induces a quiescent phenotype in these cells characterized by defective OXPHOS. These findings confirm that 2-AA mediates energy homeostatic and metabolic alterations in the immune cells and that 2-AA tolerization arrests these cells in a sustained 2-AA-unresponsive state.

The pyruvate transport into mitochondria is decreased in tolerized macrophages

Since acetyl-CoA and ATP levels were reduced due to 2-AA tolerization, we investigated whether the mitochondrial dysfunction observed is related to pyruvate metabolism. Pyruvate links glycolysis with mitochondrial production of acetyl-CoA and ATP via the tricarboxylic acid (TCA) cycle and OXPHOS [7]. RAW 264.7 cells initially exposed to 2-AA for 1 or 3 h (black bars) exhibited significantly higher concentrations of cytosolic and mitochondrial pyruvate compared to the corresponding naïve control cells (Fig. 2A). Pyruvate significantly decreased over time in cytosolic and mitochondrial fractions as cells were entering into the tolerized state (Fig. 2A). Tolerized cells remain unresponsive to 2-AA re-exposure (red bars), exhibiting no significant change in the concentration of pyruvate in the cytosolic or mitochondrial fraction (Fig. 2A). A similar result was observed in THP-1 cells (Figs. S1E and S1F).

2-AA perturbs the mitochondrial MPC1-mediated import and metabolism of pyruvate.

(A) Cytosolic and mitochondrial pyruvate levels following 2-AA exposure (black) or re-exposure (red) and corresponding controls in grey and pink respectively, for indicated time points. (B) Representative Western blot and results of densitometric analysis of MPC1 protein levels following 2-AA exposure or re-exposure for indicated time points. β-actin was used as a control. Corresponding controls are shown in grey or pink, respectively. Mean ± SD is shown, n = 3, ***p < 0.001, ****p < 0.0001, and ns indicates no significant difference. One-way ANOVA followed by Tukey’s post hoc test was applied.

To gain a better understanding of the 2-AA-mediated reductions in mitochondrial pyruvate, ATP production, and acetyl-CoA levels, we sought to determine the involvement of the MPC1 [43] that is crucial in transporting pyruvate from the cytosol into the mitochondria [44]. Western blot studies of whole cell lysates of RAW 246.7 cells exposed to 2-AA showed a significant decrease over time in the abundance of MPC1 protein, and tolerized cells remain unresponsive to 2-AA re-exposure (Fig. 2B). Together these findings suggest that 2-AA tolerization dysregulates MPC1-mediated transport of the pyruvate into mitochondria.

2-AA tolerization suppresses ERRα binding to the MPC1 promoter and the interaction between PGC-1α and ERRα

Given that the PGC-1α/ERRα axis is the transcriptional regulatory node that activates the expression of MPC1 [33] we determined the protein levels of ERRα. Western blot studies of whole cell lysates showed a significant decrease over time in the abundance of ERRα protein following 1st exposure of RAW 246.7 cells to 2-AA, while tolerized cells remain unresponsive to 2-AA re-exposure (Fig. 3A). By assessing the cytosolic and nuclear abundance of ERRα in tolerized cells we found that both fractions had lower levels of ERRα protein than their corresponding control cells (Fig. 3B).

2-AA-mediated macrophage tolerization deranges PGC-1α/ERRα-dependent metabolic programing

(A) Representative Western blot and results of densitometric analysis of ERRα protein levels following 2-AA exposure or re-exposure for indicated time points. β-actin was used as a control. (B) Western blots and its corresponding densitometric analysis of ERRα in cytoplasmic or nuclear lysates of tolerized macrophages exposed to 2-AA for 24 h or not exposed to 2-AA. (C) ChIP-qPCR assay of ERRα binding at the MPC1 promoter in RAW 264.7 tolerized macrophages exposed to 2-AA (200 μM) for 24 h (black) compared to untreated control macrophages (grey). IgG served as a negative control (D) Representative Western blot of co-immunoprecipitation (co-IP) studies of ERRα and PGC-1α in nuclear extracts of 2-AA tolerized (24 h) and control RAW 264.7 cells. Pull down with IgG served as a negative control. 2-AA tolerized macrophages are shown in black, and untreated control macrophages in grey. Mean ± SD in shown, n ≥ 3, *p < 0.05, **p < 0.01, ****p < 0.0001, and ns indicates no significant difference. One-way ANOVA followed by Tukey’s post hoc test was applied.

Furthermore, because an ERRα binding putative DNA motif (TNAAGGTCA) has been identified upstream of the MPC1 promoter in humans [43], and we found the same motif to be present 1.5 kB upstream of the transcription start site of mouse MPC1 promoter, we analyzed if tolerization led to reduced ERRα binding to the MPC1 promoter motif TNAAGGTCA. Chromatin immunoprecipitation (ChIP) assay followed by qPCR analysis revealed that in tolerized cells, ERRα binds approximately 4-fold less efficiently to the putative binding site on the MPC1 than in corresponding control cells (Fig. 3C).

Since the transcriptional activity of ERRα is regulated through a protein-protein interaction with the transcriptional coactivator PGC-1α [33] we assessed whether the formation of the PGC-1α/ERRα complex is affected in 2-AA tolerized cells. Using paraformaldehyde fixed nuclear lysates of RAW 264.7 tolerized and naïve cells, we performed a co-immunoprecipitation (co-IP) of PGC-1α and reverse co-IP with ERRα followed by immunoblotting (Fig. 3D). As shown in Figure 3D less ERRα was detected when PGC-1α was pulled down in presence of 2-AA than in naïve control cells. Moreover, reverse co-IP with ERRα protein showed lower levels of PGC-1α in the presence of 2-AA than in naïve control cells (Fig. 3D). Control IgG showed no detectable levels of ERRα following nuclear fraction pull down. These results indicate that in the tolerized macrophages, PGC-1α/ERRα interaction is impaired (Fig. 3D).

Moreover, since PGC-1α enhances the transcriptional activity of ERRα through their interaction, and ERRα transcription is regulated via an autoregulatory loop [34], we examined the effect of tolerization on the transcription of ERRα and its target gene MPC1. Using RAW 264.7 cells infected with the wild type PA strain PA14 or isogenic ΔmvfR mutant, which does not produce 2-AA [21] we observed lower RNA transcript levels of MPC1 (Fig. 4A) and ERRα (Fig. 4B) in PA14 compared to ΔmvfR. The addition of 2-AA to ΔmvfR decreased the levels of both gene transcripts to t levels similar to the PA14 infection condition, supporting the role of 2-AA in effect. To test if the observed effects are due to MPC1-dependent reduction of mitochondrial ATP generation, we supplemented the macrophages with ATP. The findings indicate that the addition of ATP in PA14 infected cells elevated the transcript levels of both ERRα and MPC1, but not in the ΔmvfR infected macrophages (Fig. 4A and 4B). These results indicate that reduced MPC1 function is due to 2-AA tolerization on the transcriptional activation of MPC1 through the PGC-1α/ERRα axis.

Increased intracellular burden in macrophages is associated with decreased expression of MPC1, ERR-α, and TNF-α genes.

Real-time PCR analysis of MPC1 (A) and ERR-α (B) expression in RAW 246.7 macrophages infected with PA14 or ΔmvfR in the presence or absence of exogenous addition of 2-AA or ATP for 6 h as indicated. Transcript levels were normalized to 18srRNA. PA14 infected cells served as controls. (C) Intracellular burden of PA14 or ΔmvfR of infected macrophages in the presence or absence of exogenous addition of 2-AA, UK5099, or ATP. Untreated cells infected with PA14 were set as 100%. D) Real-time PCR analysis of TNF-α expression in RAW 246.7 macrophages infected with PA14 or ΔmvfR in the presence or absence of exogenous addition of 2-AA, ATP, or UK5099. Transcript levels were normalized to 18srRNA. PA14 infected cells served as controls. The compound concentration used for UK5099 was 10 µM and ATP 20 µM. Mean ± SD is shown, n = 3, * p < 0.05, **p < 0.01, ***p < 0.001, ****p < 0.0001***, and ns indicates no significant difference. One-way ANOVA followed by Tukey’s post hoc test was applied.

2-AA tolerization impairs macrophage-mediated intracellular bacterial clearance through decrease in MPC1-mediated pyruvate import, ATP, and TNF-α levels

2-AA-mediates persistence of PA in vivo, dampens proinflammatory responses and increases the intracellular burden of this pathogen in macrophages via epigenetic modifications [17, 18, 26]. Here we used RAW 264.7 cells that were infected with the PA14 or the 2-AA deficient ΔmvfR mutant to assess the clearance of PA by macrophages. Macrophages infected with PA14 showed increased bacterial burden than the cells infected with ΔmvfR, and 2-AA addition to ΔmvfR led to increased bacterial burden (Fig. 4C). Furthermore, we used UK5099 and ATP to interrogate whether MPC1 mediated mitochondrial pyruvate import and bioenergetics are linked to the clearance of PA intracellular burden. The addition of the UK5099 inhibitor strongly enhanced the bacterial intracellular burden in ΔmvfR infected macrophages compared to the non-inhibited cells (Fig. 4C). UK5099 had no effect on the PA14 burden in macrophages. Conversely, exogenously added ATP to macrophages infected with PA14 strongly reduced the PA intracellular burden (Fig. 4C).

Given that 2-AA tolerization decreases the expression of pro-inflammatory cytokine TNF-α by hypo-acetylating the core histone 3 lysine 18 acetylation (H3K18ac) mark at TNF-α promoter [18], we investigated the link between bioenergetics and TNF-α expression in infected macrophages. UK5099, or ATP was added exogenously to infected macrophages with PA14 or ΔmvfR (Fig. 4D). As shown in Figure 4D, PA14 infected cells showed lower TNF-α transcript levels compared to the ΔmvfR infected cells. Supplementation of 2-AA to ΔmvfR infected cells led to a decrease in the TNF-α transcript levels. Moreover, the addition of the MPC1 inhibitor, UK5099 in ΔmvfR infected counteracted the increase in TNF-α transcript levels, while had no effect in PA14 infected macrophages (Fig. 4D). Conversely, exogeneous addition of ATP to PA14 infected cells enhanced the transcription of TNF-α compared to the untreated PA14 infected cells, while no changes in TNF-α expression levels were observed in ΔmvfR infected macrophages in presence of absence of ATP (Fig. 4D).These findings strongly suggest that 2-AA tolerization severely alters macrophages’ ability to facilitate the clearance of PA intracellular burden, via the reduction in MPC1-mediated pyruvate import into mitochondria, ATP levels and TNF-α transcription.

In vivo studies corroborate the 2-AA-mediated decrease in vitro of the central metabolic fuel acetyl-CoA and the energy-carrying molecule ATP and their association with PA persistence

To determine whether the decrease in the key metabolites mediated by 2-AA is also observed in vivo during infection, we quantified the levels of ATP and acetyl-CoA in murine spleen tissues at 1, 5, and 10 days (Fig. 5). This organ was selected for our immunometabolic studies due to its role in regulating not only local but also systemic (whole-body) immune responses, facilitated by various immune cells including macrophages [45]. Mice were infected with PA strain PA14, or isogenic mutant ΔmvfR, which does not produce 2-AA [21]. It is important to note that it is not possible to generate or utilize a bacterial mutant that is defective in 2-AA only because 2-AA is formed by spontaneous decarboxylation rather than by an enzyme-catalyzed reaction [46–48]. Therefore, animals that were either infected with ΔmvfR and received 2-AA (ΔmvfR+2-AA) at the time of infection or uninfected mice injected with a single dose of 2-AA served as direct controls (Fig. 5). Naïve and sham mice groups served as additional basal controls (Fig. 5).

2-AA promotes a long-lasting decrease in ATP and acetyl-CoA levels and bacterial persistence in PA-infected mice. (A) ATP and (B) acetyl-CoA concentrations in the spleens of mice infected with PA wild type (PA14), the isogenic mutant ΔmvfR, ΔmvfR injected with 2-AA at the time of infection (ΔmvfR + 2-AA), or non-infected but injected with 2-AA (6.75 mg/kg). (C) Bacterial burden in muscles expressed as CFU count was analyzed using the Kruskal–Wallis non-parametric test with Dunn’s post-test; ***p < 0.001, and ns indicates no significant difference. Control mice groups: naïve were not given 2-AA; mice receiving 2-AA were given a single intraperitoneal injection of 2-AA; sham represents a burn/PBS group since the burn and infection model was used. Results of three independent replicates with 4 mice per group for 1, 5, and 10 d are shown. Means ± SD are shown, *p < 0.05, **p < 0.01, ***p < 0.001, and ns indicate no significant difference. One-way ANOVA followed by Tukey’s post hoc test was applied.

Infection with PA14 or uninfected mice injected with 2-AA led to a significant decrease in both ATP (Fig. 5A) and acetyl-CoA (Fig. 5B) concentrations in spleen tissues compared to ΔmvfR infected mice that sustained higher ATP and acetyl-CoA levels similar to naïve and sham control groups across the time points tested. However, in mice infected with ΔmvfR and 2-AA injected (ΔmvfR+2-AA) at the time of infection, ATP and acetyl-CoA concentrations in the spleen decreased to levels comparable to those of PA14-infected animals, strongly indicating the biological function of 2-AA in decreasing these key metabolites. These in vivo findings further support the adverse action of 2-AA on host energy homeostasis and metabolism observed in vitro studies.

The metabolic alterations observed are associated with the host tolerance to PA persistence (Fig. 5C). Using Pseudomonas isolation agar plates, we evaluated the bacterial load at the infection site over the course of 10 days, obtaining samples at 1, 5, and 10 days. At one day post-infection, mice infected with PA14, ΔmvfR, or ΔmvfR+2-AA exhibited no difference in bacterial burden (Fig. 5C), verifying the ability of all strains to establish infection. At 5-and 10-days post-infection, mice infected with PA14 or ΔmvfR+2-AA sustained the bacterial burden at a significantly higher bacterial burden over time compared to those infected with ΔmvfR (Fig. 5C). These results are consistent with our previous findings [17, 18] and provide compelling evidence that the metabolic alterations identified may favor PA persistence in infected tissues.

Discussion

The PA signaling molecule 2-AA that is abundantly produced and secreted in human tissues [21, 22] is the first QS molecule that epigenetically reprograms immune functions, promotes immune tolerance, and sustains PA’s presence in host tissues [17, 18, 22, 26]. The present study provides insights into the mechanistic actions of 2-AA on cellular metabolism that contribute to host immune tolerance to PA persistence. We uncover that this signaling molecule causes distinct metabolic alterations in macrophages’ mitochondrial respiration and energy production promoted via the PGC-1α/ERRα axis and MPC1-dependent OXPHOS that links the TCA cycle to the production of ATP and energy homeostasis (Fig. 6). Our results show that 2-AA tolerization decreases ATP the crucial energy metabolite and the histone acetylation metabolite acetyl-CoA and in vivo, implicating the importance of the key energy-producing mitochondrial process in immune tolerance. Although macrophages respond to the first exposure to 2-AA by enhancing ATP and acetyl-CoA production, long-term exposure to 2-AA leads to a tolerized state characterized by sustained reduced ATP and acetyl-CoA concentrations, a quiescent bioenergetic state with reduced mitochondrial pyruvate uptake and unresponsiveness to a 2nd exposure.

Proposed mechanism by which 2-AA impairs bioenergetics through the inhibition of MPC1-mediated pyruvate transport into mitochondria and its impact on the PGC-1α/ERRα axis.

2-AA tolerized macrophages exhibit diminished pyruvate levels in mitochondria due to the decreased expression of MPC1, a consequence of the 2-AA impact on the interaction of ERRα with the transcriptional coactivator PGC-1α for the effective transcription of ERRα, since ERRα controls its own transcription and that of MPC1. In the presence of 2-AA, the weakened interaction between ERRα and PGC-1α results in reduced expression of MPC1 and ERRα. The reduction in mitochondrial pyruvate levels leads to decreased acetyl-CoA and ATP levels, which modulate HDAC1-and HAT-catalyzed remodeling of H3K18 acetylation. The diminished levels of this epigenetic mark have previously been associated with an increased intracellular presence of bacteria in macrophages, as demonstrated by our group [18]. Pathways, proteins, and metabolites that are negatively affected are indicated in red, while positively affected are denoted in black. The figure was created using BioRender.com.

Our results reveal that the decrease of ATP and acetyl-CoA in tolerized macrophages results from the 2-AA-mediated perturbation of the physical interaction between ERRα and PGC-1α. The PGC-1α/ERRα axis has not previously been implicated in tolerization but has been extensively studied in cancer, underscoring the novelty of our findings. Studies in cancer cells have shown that the physical interaction of human ERRα with PGC-1α strongly enhances the binding of ERRα to DNA to induce the activation of the targeted genes [33] including MPC1, responsible for the import of cytosolic pyruvate into mitochondria and PDH, that catalyzes the conversion of pyruvate to acetyl-CoA in mitochondria. Inhibition of ERRα has been shown in cancer cell metabolism studies to interfere with pyruvate entry in mitochondria by inhibiting the expression of MPC1 [49]. Indeed, we find that the transcription of ERRα and MPC1 genes and pyruvate uptake into mitochondria is reduced in tolerized macrophages. Given that pyruvate is a primary carbon source in the TCA cycle that fuels OXPHOS and the production of ATP into mitochondria, it explains the reduced ATP and acetyl-CoA levels observed in tolerized macrophages. Future studies will focus on unveiling the pathways by which macrophages respond to the bioenergetic changes identified and decipher their effect on the pyruvate cycle in host tolerance to infection [50–52].

We show that the unprecedented action of a QS bacterial small signaling molecule on the interaction between ERRα and PGC-1α in the nucleus is hampered. Although it remains to be elucidated whether 2-AA interferes directly or indirectly with the PGC-1α/ERRα axis, our data bring up the possibility that the reduced production of acetyl-CoA may be responsible, since cholesterol is a known coactivator/ligand of ERRα [53], is generated from acetyl-CoA [53]. This possibility may also explain the reduced expression of ERRα and its lower abundance since cholesterol is also required for its activation and autoregulation. However, it is also possible that 2-AA may antagonize the interaction between ERRα and PGC-1α by binding to ERRα. Small molecules as antagonist of ERRα have been reported previously [54]. The weaker binding of ERRα to the MPC1 promoter site, and consequently reduced MPC1 expression explains the reduced pyruvate presence in mitochondria and that of ATP and mitochondrial quiescence.

The observed cellular metabolic perturbances also align with our previous studies in skeletal muscle, which pointed to mitochondrial dysfunction triggered by 2-AA [24, 25]. Interestingly, the PA QS LasR-regulated homoserine lactone molecule, 3-oxo-C12-HSL, was reported to attenuate the expression of PGC-1α and decrease the mitochondrial respiratory capacity in lung epithelial cells [27]. Moreover, another PA QS molecule PQS also regulated by LasR was recently shown to induce organelle stress, including mitochondria, by disrupting the mitochondrial membrane potential in human macrophages [55]. As opposed to 2-AA, which dampens the pro-inflammatory response, PQS increases proinflammatory cytokines [55], consistent with the fact that PQS is an acute infection type molecule unrelated to chronic/persistent infections.

Previously, we reported that 2-AA tolerization induces HDAC1 expression and activity, reducing H3K18ac at the TNF-α promoter [18]. The findings with TNF-α transcription using UK5099 and ATP in 2-AA treated macrophages are in support of the bioenergetics disturbances observed in macrophages and their link to epigenetic modifications we have shown to be promoted by 2-AA [18]. Macrophages exposed to 2-AA in the presence of exogenous ATP showed improved intracellular bacterial clearance and enhanced TNF-α levels, supporting the epigenetic interconnection of macrophages and cellular ATP responsiveness levels against infection. The exogenous addition of UK5099 that reverted the efficiency of the mvfR infected macrophages to clear the PA intracellular burden and the counteraction to the 2-AA effect by ATP addition that increases the clearance of PA intracellular burden support the importance of this energy-carrying molecule in PA persistence.

The results with the MPC1 inhibitor, UK5099, suggest that the availability of pyruvate could underlie the mechanism of 2-AA-regulated HDAC/HAT-dependent control of transcription of pro-inflammatory mediators and bacterial clearance [56]. These results align with our previous findings, showing that establishing bacterial intracellular burden in macrophages is HDAC1 dependent [56]. It remains to be elucidated if HDAC/HAT-mediated histone modification directly regulates expression of OXPHOS genes in 2-AA tolerized macrophages as it was shown that histone deacetylation downregulates OXPHOS in persistent M. tuberculosis infection [57, 58]. Given the complexity of 2-AA-mediated long-term effects, future omics studies combined with immune profiling may aid in deciphering the possible network of co-factors across different subpopulations of immune cells and the immunometabolic reprogramming related to 2-AA.

We have shown that although 2-AA tolerization leads to a remarkable increase in the survival rate of infected mice, it permits an HDAC1-dependent sustained presence of PA in mice tissues and intracellularly in macrophages [17, 18, 22, 26]. Here, we use 2-AA-producing and isogenic non-producing PA strains to confirm the 2-AA-mediated decrease in the central metabolite acetyl-CoA and the energy-carrying molecule ATP during infection. These murine infection studies provide strong evidence of the 2-AA biological relevance in reducing these metabolites in vivo. Taking together our previous and current in vivo findings provide compelling evidence that the metabolic alterations identified favor PA persistence in infected tissues.

That 2-AA permits PA to persist in infected tissues despite rescuing the survival of infected mice underscores its difference from that of LPS tolerization, which results in bacterial clearance [23] via the mechanism that relies on a set of different HDAC enzymes. LPS tolerization predominantly involves changes in H3K27 acetylation [59], while 2-AA tolerization involves H3K18 modifications [22]. The 2-AA-mediated effects reported here are also distinct from the epigenetic-metabolic reprogramming mediated by LPS [60], which promotes endotoxin tolerance in immune cells by upregulating glycolysis and suppressing OXPHOS [61]. Although 2-AA and LPS implicate different components and lead to different outcomes both involve epigenetic mechanisms and immune memory.

Metabolic reprogramming upon infection may be pathogen-specific, with each pathogen impacting specific metabolic pathways that better fit its respective metabolic needs [7]. This was shown for infections of various tissues with Chlamydia pneumonia, Legionella pneumophila, Mycobacterium tuberculosis, and Salmonella [62–65] and, by our group, for PA infections [17, 18, 24]. It would be interesting, however, to test whether 2-AA affects the killing efficacy of macrophages against other pathogens, as emerging evidence shows that synergistic or antagonistic interactions between clinically relevant microorganisms and host have important implications for polymicrobial infectious diseases [66].

While in this study, we focused on the role of ERRα mainly in pyruvate metabolism, future studies are needed to reveal other ERRα/PGC-1α axis-dependent metabolic and immune pathways are modulated by 2-AA. In this view, the cholesterol synthesis pathway that relies on acetyl-CoA precursor, as cholesterol is a known coactivator/ligand of ERRα [53] and fatty acid catabolism, generating acetyl-CoA, and PGC-1α regulates fatty acid oxidation [67].

This study unveils the unprecedented actions of a QS bacterial molecule in orchestrating cellular metabolic reprogramming in addition to the epigenetic reprogramming reported previously also shown to promote a long-lasting presence of PA in the host [17, 18, 26]. That 2-AA cross-tolerized macrophages to LPS corroborates our previous findings on the implication of epigenetic mechanism [18] rather than ligand-receptor mediated signaling-based mechanism, and raises the possibility that this QS molecule may confer non-specific cross protection. An aspect we plan to investigate in the future. The reported immunometabolic reprogramming contributes to a better understanding of the molecular and cellular mechanisms, biomarkers, and functional significance that may be involved in immune tolerance to persistent infection, providing for designing and developing innovative therapeutics and interventions. These approaches can focus on promoting host resilience against bacterial burden and safeguarding patients from recalcitrant persistent infections.

Materials & Methods

Ethics statement

The animal protocol was approved by the Institutional Animal-Care and Use Committee (IACUC) of Massachusetts General Hospital (protocol no. 2006N000093). No randomization or exclusion of data points was applied.

Conditions used for the exposures of the cells to 2-AA

Mouse BMDM cells were isolated from femur of CD1 6 weeks old male and female mice and used to estimate the levels of various metabolites (Fig. 2A) following exposure and re-exposure to 2-AA (Fig. 1A). The 1st exposure of naïve BMDMs (106/mL in 6-well plates) was achieved with 400 µM 2-AA for 1, 6, or 48 h. Cells exposed for 48 h were used for the 2nd 2-AA exposure. These cells were washed to remove residual 2-AA, left in the medium in the absence of 2-AA for either 24 or 106 h, and re-exposed to 2-AA (200 µM) for 1 or 6 h, respectively. For RAW 264.7 (106/mL in 6-well plates) and THP-1 cells (106/mL in 6-well plates) the 2-AA concentration and conditions used were as with BMDM cells. The hours of the 1st round and 2nd round of exposure to 2-AA are indicated in Figure S1.

For the LPS stimulation studies, cells were stimulated with 400 µM 2-AA or 100 ng/mL LPS for 6 h, and 800 µM of 2-AA for 48 h. Cells receiving 1st exposure for 48 h were washed and re-exposed to 400 µM of 2-AA or 100 ng/mL LPS for 6 h.

ATP and acetyl-CoA quantifications

The levels of ATP, acetyl-CoA, and were assessed in BMDM and RAW264.7 macrophage cells following exposure to 2-AA at the times indicated and by utilizing the ATP Assay kit (cholorimetric/fluorometric) (#ab83355, Abcam) and the PicoProbe AcCoA assay kit (ab87546, Abcam), respectively, according to the manufacturer’s instruction. Quantifications of ATP and acetyl-CoA were performed in triplicates.

ATP: For ATP determination, macrophages were lysed and subsequently centrifuged at 14,000 rpm for 10 min at 4 °C. The supernatants were transferred to a fresh Eppendorf tube. Standards and cell supernatants of 50 µL were added to a 96-well plate suitable for fluorescent analysis (black sides, clear bottom). A reaction mixture containing an ATP converter, probe, buffer, and developer mix was then added to all wells (50 µL) and incubated away from light at room temperature for 30 min. ATP quantification was conducted fluorometrically at 535/587 nm using a plate reader and the following settings: λex = 535 nm; λem = 587 nm. Fluorescence was measured using a microplate reader (Tecan Group Ltd, Männedorf, Switzerland)

Acetyl-CoA: Acetyl-CoA content was assessed by first deproteinizing total cell fractions of macrophages using the perchloric acid and then centrifuging at 14,000 rpm for 10 min at 4 °C. 50 µL of cells’ supernatant sample CoASH were quenched to correct the background generated by free CoASH. Following the homogenization procedure described above, the samples were diluted with the reaction mix, and fluorescence was quantified using a plate reader and the settings as above with ATP.

Seahorse Assays

Seahorse analysis was performed according to previously published protocols [68]. Briefly, freshly prepared BMDM cells were reseeded in complete RPMI-1640 cells using Seahorse plates (Agilent cat. no. 103729100) at a density of 5 × 104 cells per well. Cells were exposed to 400 µM 2-AA for 1 h or 24 hours (1st exposure). Cells receiving 1st exposure for 24 h were washed and re-exposed to 200 µM of 2-AA for 1 h. Naïve cells were used as control. Prior to initiating Seahorse measurements cells were washed and Seahorse XF DMEM supplemented with 2 mM glutamine (Gibco cat. No. 25030-081) was added to each well. Cells were allowed to stabilize in a 37°C incubator without CO2 for 1 h. The Seahorse cartridge was hydrated and calibrated as per the manufacturer’s instructions. The Mitochondrial stress test kit (Agilent cat. no. 10395-100) was used (oligomycin at 1 μM, FCCP (1.5 µM), rotenone (0.5 μM), and antimycin A at (0.5 μM)) with slight modifications according to the published protocol [68] that included injection of 25 mM glucose, and sodium pyruvate (1 μM). All samples N= 4 were run in a Seahorse XFe96 Analyzer, and data were analyzed using Wave and plotted using GraphPad Prism software.

Isolation of cytosolic and mitochondrial fraction and pyruvate quantification

Pyruvate is produced in the cytosol and is transported into the mitochondria. RAW 264.7 cells were plated at 1 × 106 to incubate overnight at 37°C in a CO2 incubator. Cells were exposed to 2-AA for 1, 3, 24 or 48 h or re-exposed for 1 h as described above. Mitochondria and cytosolic fractions of each group were isolated utilizing the Mitochondria Isolation Kit for cultured cells (#ab110170, Abcam) following the manufacturer’s protocol. Briefly, cells were collected with a cell lifter and pelleted by centrifugation at 1,000 g, frozen, and then thawed to weaken the cell membranes. The cells were resuspended in Reagent A and transferred into a pre-cooled Dounce Homogenizer. The homogenates were centrifuged at 1,000 g for 10 min at 4°C and saved as supernatant #1 for the cytosolic fractions. The pellet was resuspended in Reagent B, followed by repeat rupturing and centrifugation. The pellet was collected and resuspended in 500 μL of Reagent C supplemented with Protease Inhibitor cocktails (P8340, Sigma Aldrich). Following separation, a Bradford assay was conducted to determine the protein concentration in each fraction. Pyruvate levels were determined in cellular fractions and mitochondrial fractions by using the Pyruvate Assay Kit (#ab65342, Abcam) according to the manufacturer’s instructions. Briefly, after deproteinization using perchloric acid, the samples were neutralized in ice-cold 2 M KOH. 10 µL samples were incubated with reaction mix and kept on the plate at RT for 30 min in the dark. The absorbance was measured in a microplate reader (Tecan Group Ltd, Männedorf, Switzerland) at 570 nm and results are shown from three independent experiments.

Bacterial strains and growth conditions

The PA strain known as RifR human clinical isolate UCBPP-PA14 (also known as PA14) was used [69]. The bacteria were grown at 37 °C in lysogeny broth (LB) under shaking and aeration or on LB agar plates containing appropriate antibiotics. PA14 and isogenic mutant ΔmvfR [19] cultures were grown in LB from a single colony to an optical density of 600 nm (OD600) of 1.5, diluted 1:50,000,000 in fresh LB media, and grown overnight to an OD600 of 3.0.

Gentamicin protection assay

RAW264.7 macrophage cells were plated on 6 well cell culture-treated plates overnight in pyruvate free DMEM. After 3 h of incubation with 10 µM UK5099, 20 µM ATP or 400 µM 2-AA, cells were infected with PA14 and isogenic mutant ΔmvfR at 5 MOI for 30 min at 37 °C in 5% CO2. Unbound bacteria were removed by washing once with cold DMEM medium. Afterwards, the cells were incubated with 100 µg/ml of gentamicin for 30 min to eliminate residual extracellular bacteria. The cells were washed with DMEM and transferred to free medium without gentamicin and kept for 3 h at 37°C in 5% CO2. The infected cells were scraped after 3 h, centrifuged at 500 × g, and lysed in distilled water. The lysed cells were immediately diluted in PBS and plated on LB agar plates to assess bacterial presence. Bacterial colony-forming units (CFUs) were counted after incubating the plates overnight at 37°C.Untreated cells infected with PA14 were set as 100 % and the reduction of Bacterial load was expressed as %CFU.

Pharmacological inhibitors and ATP supplementations

For the OXPHOS inhibition assay, RAW264.7 macrophage cells were treated with UK5099 (10 µM, Sigma-Aldrich, dissolved in DMSO 3 h prior to 1st or 2nd rounds of 2-AA exposure. RAW264.7 macrophages supplemented with ATP (20 µM, Sigma-Aldrich, dissolved in PBS) received ATP at the time of 1st and 2nd 2-AA exposure.

Co-Immunoprecipitation assay

For all immunoprecipitation assays, protein A/G agarose Magnetic beads (Pierce) were used after washing in 1X IP buffer. Nuclear lysates from RAW 264.7 tolerized cells exposed to 2-AA for 24 h or non-exposed cells were diluted in 1X IP buffer and 500 µg of protein was taken for each experiment. The lysates were precleared by using unbound 50 µl protein A/G magnetic beads for 2 h at room temperature on a rotator. Precleared lysates were either incubated with ERRα, PGC-1α or rabbit IgG antibody overnight at 4 °C. 100 µl protein A/G magnetic beads were used to pull down the antibody-protein complex and washed twice with IP buffer to remove unbound proteins. Magnetic beads were then eluted in 2X Laemmli SDS-PAGE loading buffer. Immunoprecipitated ERRα and PGC-1α was detected by Western blot analyses, using conformational specific Anti Rabbit antibody (TruBlot).

Immunoblotting analysis

Cells were plated at 6×105 cells per well in six-well plates. Cells were washed with PBS and subsequently lysed using RIPA lysis buffer containing 1 mM phenylmethylsulfonyl fluoride (PMSF). 20μg of proteins were separated by electrophoresis on any KD (Kilo Dalton) (Bio-Rad, Cat no. 4569033) SDS-polyacrylamide gel. Proteins were transferred to a 0.2 μm polyvinylidene fluoride membrane (PVDF, Millipore, Billerica, USA) using a Bio-Rad semi-dry instrument. After blocking with 5% BSA in TBS containing 0.1% Tween-20 for 1 h at room temperature, the membranes were incubated with a primary antibody ERR-a (Abcam, #ab76228), MPC1 (D2L91, Cell signaling), and anti-β-actin (cat no: sc-47778) (Santa Cruz Biotechnology) overnight at 4°C. Following washing, the membranes were incubated with an anti-rabbit secondary antibody, and the bands were detected by SuperSignal West Pico Chemiluminescent Substrate (Thermo Scientific) reaction, according to the manufacturer’s instructions. The blots were visualized in the ChemiDOC Imaging system (Bio-Rad Laboratories, Inc., Hercules, CA, USA). The bands were analyzed densitometrically using QuantityOne software (Bio-Rad).

Chromatin immunoprecipitation (ChIP) and ChIP-qPCR

For ChIP studies, RAW264.7 macrophage cells exposed to 2-AA for 24 h were cross-linked in 1% (v/v) methanol-free formaldehyde for 10 min and then placed in 0.125 M glycine for 5 min at RT. Using the truChIP High Cell Chromatin Shearing kit (Covaris, United States), cells were prepared for sonication according to the Covaris protocol. Approximately 1 × 107 cells were plated in a 12 mm × 12 mm tube and subjected to shearing with the Covaris S220 sonicator for 8 min (140 peak power, 5 duty factor, 200 cycles/burst). The Magna ChIP A/G kit (Millipore, United States) was used for the subsequent immunoprecipitations according to the manufacturer’s protocol. Briefly, chromatin from approximately 106 cells was incubated overnight at 4°C with 2 μg of anti-ERRα or anti-PGC-1α (Abcam, USA) ChIP-grade antibody and 20 μL of A/G magnetic beads. The beads were washed serially (5 min each) with low-salt wash buffer, high-salt wash buffer, LiCl wash buffer, and TE buffer from the kit at 4°C. Chromatin was eluted with elution buffer containing Proteinase K at 62°C for 4 h, then incubated at 95°C for 10 min. DNA was isolated by column purification (QIAquick PCR purification kit).

Real-time ChIP-qPCR was performed with the Brilliant II SYBR green super mix (Agilent, USA). Forward (AGTGGTGACCTTGAACTTCCC) and reverse (CTGAAGACGACCTTCCCCTT) primers were chosen to amplify a genomic locus of MPC1 promoter which had a putative ERR binding site (TNAAGGTCA) at 1582 bp upstream of the start site. Normalized values were calculated by the percent-input method relative to the IgG. The assay was performed three times.

RNA extraction and RT-qPCR

Total RNA from all the groups mentioned above was isolated from approximately 2 × 106 cells with the RNeasy minikit (Qiagen, USA), and cDNA was prepared with the iScript Reverse transcription kit (Bio-Rad, USA), as per the manufacturer’s instruction. Real-time PCR was conducted using the PowerUP SYBR green Master mix (Applied Biosystem, USA) and primer sets for mouse ERR-α (forward: ACTACGGTGTGGCATCCTGTGA; reverse: GGTGATCTCACACTCATTGGAGG), MPC-1 (forward: CTCCAGAGATTATCAGTGGGCG; reverse: GAGCTACTTCGTTTGTTACATGGC), TNF-α (forward: GGTGCCTATGTCTCAGCCTCTT; reverse: GCCATAGAACTGATGAGAGGGAG) and mouse 18S rRNA (forward: GTTCCGACCATAAACGATGCC; reverse: TGGTGGTGCCCTTCCGTCAAT). The transcript levels of all the genes were normalized to 18S rRNA with the ΔΔCT method. The relative expression was calculated by normalizing transcript levels to those of PA14 Infected cells. The assay was conducted in triplicate; means and standard deviations were calculated for each group.

Animal infection and metabolites assessment experiments

The full-thickness thermal burn injury and infection (BI) model [17, 18, 70] was used to assess the effect of 2-AA on metabolic alterations in the spleens of six-week-old CD1 male mice and bacterial burden (Charles River Labs, USA). A full-thickness thermal burn injury involving 5-8% of the total body surface area was produced on the shaved mouse abdomen dermis, and an inoculum of ∼1 × 104 PA14 and ΔmvfR [19] cells in 100 μL of MgSO4 (10 mM) was injected intradermally into the burn eschar. For the groups that received 2-AA, mice were injected intraperitoneally (IP) with 100 μL of 2-AA (6.75 mg/kg). The entire procedure was done under the influence of anesthesia. One of the groups of mice infected with the PA14 isogenic mutant ΔmvfR [19] also received 100 μL of 2-AA in PBS (6.75 mg/kg) at the time of infection (ΔmvfR +2-AA) and served as an additional control. CFU counts from rectus abdominus muscle (underlying the burn-infected tissue) were assessed in groups of 4 mice each at 1, 5, and 10 d post-BI by plating diluted muscle homogenate on Pseudomonas isolation agar (Sigma-Aldrich) plates containing rifampicin (50 mg/L). Spleen samples from mice were collected for metabolite analyses from all mice groups at 1, 5, and 10 d post-BI to assess ATP (#ab83355 Abcam) and acetyl-CoA (#ab87546, Abcam) as described above. Tissues were homogenized, and 100 mg of homogenate was centrifuged at 4°C for 15 min at 10,000 g, and supernatants were collected. The supernatant was mixed with 400 µL of 1 M perchloric acid. The deproteinized supernatant was neutralized by 3 M KHCO3. The ATP and acetyl-CoA assays were performed as described in the in vitro section.

eLife Assessment

This important study demonstrates that the Pseudomonas aeruginosa-derived quorum sensing signal, 2-aminoacetophenone, induces immune tolerization in macrophages by perturbing metabolism, particularly in the context of mitochondrial respiration and bioenergetics. The authors present convincing evidence for 2-aminoacetophenone-mediated reduction of pyruvate transport into mitochondria, with downstream effects that result in reduced ATP production in tolerized macrophages. The work will be of interest to those studying host-pathogen interactions.

Reviewer #1 (Public Review):

Summary:

Their findings elucidate the mechanisms underlying 2-AA-mediated reduction of pyruvate transport into mitochondria, which impairs the interaction between ERRα and PGC1α, consequently suppressing MPC1 expression and reducing ATP production in tolerized macrophages. While the data presented is intriguing and the paper is well-written, there are several points that warrant consideration. The authors should enhance the clarity, relevance, and impact of their study.

Strengths:

This paper presents a novel discovery regarding the mechanisms through which PA regulates the bioenergetics of tolerized macrophages.

Weaknesses:

The relevance of the in vivo model to support the conclusions is questionable. Further clarification is needed on this point.

Reviewer #2 (Public Review):

Summary:

The study tries to connect energy metabolism with immune tolerance during bacterial infection. The mechanism details the role of pyruvate transporter expression via ERRalpha-PGC1 axis, resulting in pro-inflammatory TNF alpha signalling responsible for acquired infection tolerance.

Strengths:

Overall, the study is an excellent addition to the role of energy metabolism during bacterial infection. The mechanism-based approach in dissecting the roles of metabolic coactivator, transcription factor, mitochondrial transporter, and pro-inflammatory cytokine during acquired tolerance towards infections indicates a detailed and well-written study. The in vivo studies in mice nicely corroborate with the cell line-based data, indicating the requirement for further studies in human infections with another bacterial model system.

Weaknesses:

The authors have involved various mechanisms to justify their findings. However, they have missed out on certain aspects which connect the mechanism throughout the paper. For example, they measured ATP and acetyl COA production linked with bacterial re-exposures and added various targets like MCP1, EER alpha, PGC1 alpha and TNF alpha. However, they skipped PGC1 alpha levels, ATP and acetyl COA in various parts of the paper. Including the details would make the work more comprehensive.

The use of public data sets to support their claim on immune tolerance is missing. Including various data sets of similar studies will strengthen the findings independently.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

We thank the reviewer for the time and effort in reviewing our revised manuscript and are grateful for their constructive comments and for acknowledging the significance of our work.

Summary:

Their findings elucidate the mechanisms underlying 2-AA-mediated reduction of pyruvate transport into mitochondria, which impairs the interaction between ERRα and PGC1α, consequently suppressing MPC1 expression and reducing ATP production in tolerized macrophages. While the data presented is intriguing and the paper is well-written, there are several points that warrant consideration. The authors should enhance the clarity, relevance, and impact of their study.

Strengths:

This paper presents a novel discovery regarding the mechanisms through which PA regulates the bioenergetics of tolerized macrophages.

Weaknesses:

The relevance of the in vivo model to support the conclusions is questionable. Further clarification is needed on this point.

We appreciate the reviewer’s comment. Our conclusion that 2-AA decreases bioenergetics while sustains bacterial burden is further supported by additional in vivo data we present now in Fig. S5. To strengthen the relevance of our in vivo data, we performed additional in vivo experiments. In this set of in vivo studies, mice received the first exposure to 2-AA by injecting 2-AA only and the 2nd exposure through infection with PA14 or ΔmvfR four days post-2-AA injection. As shown in the supplementary Figure S5 the levels of ATP and acetyl-CoA in the spleen of infected animals and the enumeration of the bacterial counts were the similar between PA14 or ΔmvfR receiving the 1st 2-AA exposure and agree with the “one-shot infection” findings presented in Figure 5 with the PA14 or ΔmvfR+2-AA infected mice or those receiving 2-AA only. These results are consistent with our previous findings showing that 2-AA impedes the clearance of PA14 (Bandyopadhaya et al. 2012; Bandyopadhaya et al. 2016; Tzika et al. 2013) and provide compelling evidence that the metabolic alterations identified may favor PA persistence in infected tissues.

Reviewer #2 (Public Review):

We thank the reviewer for the time and effort in reviewing our revised manuscript and are grateful for their constructive comments and for acknowledging the significance of our work.

Summary:

The study tries to connect energy metabolism with immune tolerance during bacterial infection. The mechanism details the role of pyruvate transporter expression via ERRalpha-PGC1 axis, resulting in pro-inflammatory TNF alpha signalling responsible for acquired infection tolerance.

Strengths:

Overall, the study is an excellent addition to the role of energy metabolism during bacterial infection. The mechanism-based approach in dissecting the roles of metabolic coactivator, transcription factor, mitochondrial transporter, and pro-inflammatory cytokine during acquired tolerance towards infections indicates a detailed and well-written study. The in vivo studies in mice nicely corroborate with the cell line-based data, indicating the requirement for further studies in human infections with another bacterial model system.

Weaknesses:

The authors have involved various mechanisms to justify their findings. However, they have missed out on certain aspects which connect the mechanism throughout the paper. For example, they measured ATP and acetyl COA production linked with bacterial re-exposures and added various targets like MCP1, EER alpha, PGC1 alpha, and TNF alpha. However, they skipped PGC1 alpha levels, ATP and acetyl COA in various parts of the paper. Including the details would make the work more comprehensive.

We appreciate the reviewer’s comments and apologize for omitting the PGC-1α levels. Per the reviewer’s suggestion, we have added the PGC-1α transcript levels (Figure 4C) in the section describing 2-AA-mediated dysregulation of the ERRα and MPC1 transcription (lines 243-252). Moreover, we have added Figure S5, which shows additional ATP and acetyl CoA levels in vivo. In our view, ATP and acetyl-CoA levels are shown in all appropriate settings, interrogating the bioenergetics, including in the presence of bacteria and in their absence, where only 2-AA is added. Please see Figures 1 and 5 and the newly added Figure S5.

The use of public data sets to support their claim on immune tolerance is missing. Including various data sets of similar studies will strengthen the findings independently.

Suppose we understand correctly the reviewer’s comment regarding public data sets on immune tolerance. In that case, we are referring to our data since there are no published data from other groups on 2-AA tolerization and because the outcome of the 2-AA effect on the bacterial burden differs from that of LPS. Therefore, this study did not consider comparing with published data from LPS.

Reviewer #1 (Recommendations For The Authors):

(1) Animal model: The authors appropriately initiated the study with an in vitro tolerization model involving 2-AA re-exposure, providing foundational insights for further investigation. However, the rationale for the one-shot injection in the in vivo model lacks clarity. To strengthen the relevance of the in vivo data, the authors should consider establishing a model involving bacterial re-exposure, such as a two-challenge paradigm with antibiotic treatment in between. This approach would allow for the examination of peritoneal macrophages harvested from mice, assessing ATP levels, acetyl CoA, TNF production, and bacterial counts. Such an approach would better align the in vivo findings with the in vitro experiments, confirming the role of tolerized macrophages in controlling PA infection in the presence of 2-AA.

We thank the reviewer for this comment. Indeed, we have performed a similar two-challenge paradigm study in which first exposure to 2-AA is achieved by injecting 2-AA, and 2nd exposure through infection with PA14 or ΔmvfR four days post -2-AA injection. The results of Figure S5 can be directly compared with those in Fig 5 in vivo studies. As shown in supplementary Figure S5 the levels of ATP and acetyl-CoA in the spleen of infected animals and the enumeration of the bacterial counts agree with the “one-shot infection” presented in Fig 5 (PA14 or ΔmvfR+2-AA). Figure S5 study although not included initially to simplify data presentation, it was performed in parallel with Fig 5 and thus they can be directly compared.

(2) Exogenous ATP treatment: It is crucial to explore whether 2-AA re-exposure suppresses inflammasome activation and whether this suppression can be reversed by exogenous ATP treatment. Specifically, the authors should investigate whether NLRP3 inflammasome activation is inhibited in tolerized macrophages and whether such activation is necessary for host defense. Clarifying these points would provide valuable insights into the mechanisms underlying macrophage tolerization induced by 2-AA.

Excellent point. We agree, indeed, this is planned in the near future.

(3) Figures 4C and D: The authors should exercise care in describing these figures. For instance, line 263 states that "UK5099 had no effect on the PA14 burden in macrophages," which requires correction for accuracy.

We apologize and rephrase this sentence and other sentences referring to Fig 4D and 4E in this section. Please see the highlighted sentences in the results section referring to Fig 4. For example, “The addition of the UK5099 inhibitor strongly enhanced the bacterial intracellular burden in ΔmvfR infected macrophages compared to the non-inhibited ΔmvfR infected cells, reaching a similar burden to those infected with PA14 (Fig. 4D)”.

(4) ERRα expression: While the study intriguingly demonstrates a decrease in ERRα levels in tolerized macrophages following exposure to 2-AA, the discussion of this finding is lacking. It is worth exploring the possibility of increasing ERRα expression to counteract the tolerization induced by 2-AA and enhance clearance of PA infection. This avenue should be thoroughly discussed in the manuscript's Discussion section, offering insights into potential therapeutic strategies to mitigate the effects of 2-AA on macrophage function.

Thank you so much for this additional comment. We have now included this point in the discussion section (lines 373-376).

Reviewer #2 (Recommendations For The Authors):

Overall, the study is an excellent addition to the role of energy metabolism during bacterial infection. The mechanism-based approach in dissecting the roles of metabolic coactivator, transcription factor, mitochondrial transporter, and pro-inflammatory cytokine during acquired tolerance indicates a detailed and well-written study. However, connecting the mechanisms often was not reflected in some of the experiments, and answering a few concerns/suggestions will undoubtedly improve the study's readability, appeal, and overall impact on a broader audience.

(1) The authors should rephrase the title if possible. The title indicates 2AA as a bacterial quorum sensing signal; however, throughout the manuscript, there are no studies associated with actual quorum sensing in bacteria.

Thank you for this comment. However, the title indicates 2-AA as a quorum sensing molecule because the synthesis of this signaling molecule is uniquely regulated by quorum sensing. Because of its importance in the virulence of Pseudomonas aeruginosa and its regulation by quorum sensing, we feel that it is appropriate to refer to it as such.

(2) The authors generalised immunotolerance and memory of 2AA-exposed cells to broad-spectrum microbial exposure by just testing with LPS exposure. I would suggest they test at least 2 more heterologous microbial products known to illicit response and confirm their claim from Figure 1.

We appreciate the reviewer’s comment. We intend not to generalize immunotolerance and memory of 2-AA exposed cells to broad-spectrum microbial exposure. Moreover, since the manuscript is not focused on comparing other bacterial molecules to 2-AA and multiple studies have focused on LPS tolerance, we tested LPS only in the manuscript.

(3) LPS triggers ATP production through glycolysis in nitric oxide (NO) dependent mechanisms in various immune and non-immune cells. The authors should study the concentrations of NO, Glucose, and Pyruvate levels to clarify the mechanism of energy dynamics and the source of ATP and Acetyl CoA generated/scavenged during primary and secondary exposures to both 2AA and LPS.

We agree that a cross-tolerization experiment using 2-AA and LPS would reveal interesting insights into immune response during PA infections. However, this is out of the scope of this article. Please notice that the mechanism of 2-AA and LPS tolerization is mechanistically distinct, e.g. they rely on different HDAC enzymes, and LPS tolerization predominantly involves changes in H3K27 acetylation (Lauterbach et al. 2019). In contrast, 2-AA tolerization involves H3K18 modifications (Bandyopadhaya, Tsurumi, and Rahme 2017). For this reason, the complexity of such interactions would require a comprehensive set of experiments that are not part of the focus of this study.

(4) Immunogenic triggers often rapidly alter mitochondrial membrane potential, which alters oxygen consumption rates. However, the authors tend to generalize energy homeostasis and claim the deregulation of OXPHOS-inducing quiescent phenotype depending upon OCR measurements from Figure 1D. The authors must evaluate mitochondrial health and membrane potential during first and second exposure in a time-dependent manner to strengthen their theory of mitochondrial dysfunction. The authors should also check the phenomena in vivo (mice exposed to infection) if possible.

Thank you for this suggestion. We now include electron microscopy images of mitochondria isolated from macrophages exposed to 2-AA. Results revealed that 2-AA alters mitochondrial morphology and cristae, supporting the mitochondrial dysfunctionality caused by 2-AA. These results are shown in Figure S4 and lines 185-188.

(5) Since both MCP1 and MCP2 transporters are known to transport pyruvate to mitochondria, checking both MCP1 and 2 at transcript and protein levels in exposed cells will be essential. I suggest authors use MCP inhibitors or use RNA interference against MCPs to check the effect on tolerance of the cells exposed for a second time.

To our understanding, mitochondrial pyruvate carrier proteins, MPC1 and MPC2, form a hetero-oligomeric complex in the inner mitochondrial membrane to facilitate pyruvate import into mitochondria (McCommis and Finck 2015). We also used UK5099 an MPC carrier inhibitor for enumeration of bacterial load in macrophages in Figure 4 and observed a similar effect as 2-AA suggesting a similar mechanism of action.

(6) The pyruvate levels of mitochondria in Figure 2A are shallow, and the authors claim statistical significance within a 1.5-fold change. The authors should cross-check the number of mitochondria they are isolating while estimating pyruvate from only mitochondrial fractions. Another point is, correlating mitochondrial pyruvate with the burst of ATP during first exposure in comparison to second exposure, one can argue that the number of mitochondria is variable between the exposures leading to a change in pyruvate amount (mitochondria number increases to compensate for the first exposure and decreases quickly to maintain homeostasis and remains quiescent during a second exposure due to activation of compensatory immune mechanism towards primary exposure). How do authors address the issue?

Our electron microscopic studies indicate that although after 2-AA exposure, no reduction in mitochondrial numbers is observed in macrophages, alterations in mitochondrial morphology and cristae are observed. Please also see our answer to point # 4.

(7) The authors claim that ERR alpha regulates MCP1 transcription via activation of ERRalpha-PGC1 alpha axis and tolerization in cells to second exposure is due to impairment of the axis (Figure 3). PGC1 alpha is known to be induced during various metabolic, physiological, and immune-challenge-related stress in a tissue-dependent manner. In this context, one should expect changes in transcript and protein levels of PGC1 alpha. The authors must study PGC1 alpha levels with time-dependent exposures. LPS was shown to induce oscillations in PGC1 alpha levels in a tissue-specific manner. In experiments, authors should verify if such oscillations persist during time-dependent exposure, emphasising mitochondrial uncoupling that might get dampened during re-exposures to microbial challenges.

We appreciate the suggestion. We have now included PGC-1α (Figure 4C) transcript levels, which show the same profile as the transcript levels of ERRα and MPC1. Please note that PGC-1α is only one of several ERRα co-activators; therefore, the amount of ERRα protein is the most relevant assessment regarding the activation of the MPC1 transcription.

(8) The authors claim that ERRalpha induces MCP1 through ChIP data in Figure 3. However, the physical verifications at mRNA levels and mutational/inhibitor-based experiments are missing. The authors should study the alterations of MCP1 mRNA in relation to exposures and inhibitors of ERRalpha and PGC1 alpha to strengthen their work.

This is an interesting approach; however, this experiment exceeds the scope of our manuscript. We will certainly consider this suggestion in our future experiments. Thank you.

(9) Publicly available data sets with LPS exposures should be analyzed for gene sets pertaining to mitochondrial OXPHOS, metabolism, immune response, etc. This will support the authors' work and provide a global overview of transcriptome associated with immune tolerance.

We appreciate the reviewer’s comment. For the reasons explained in #3 point and because the bacterial burden outcome of the 2-AA effect is different from that of LPS, comparison with LPS published data was not considered in this study. We agree that in the future, a comprehensive comparison of whole genome transcriptome studies between LPS and 2-AA may reveal important insights that may also help better understand and potentially classify the immune tolerance triggered by 2-AA.

(10) In Figure 4, the authors study the role of MCP1 and associated pyruvate-dependent bacterial clearance during tolerization and associate them with a decrease in TNF alpha. I would suggest the addition of an ERR alpha inhibitor in these experiments. It is not clear as to why (mechanism) TNF alpha transcription was affected via pyruvate transport during bacterial exposure. I would suggest that the authors clarify the mechanism of TNF alpha activation/inactivation and its association with energy metabolism during acquired tolerance.

This is an excellent suggestion, given that a similar effect of ERRα on TNF-α was observed by other researchers (Chaltel-Lima et al. 2023). Here, to clarify the mechanism of TNF alpha activation/inactivation and its association with energy metabolism, we elaborate on this aspect in the discussion section.

Lines 388-393. The text reads:

Previously, we reported that 2-AA tolerization induces histone deacetylation via HDAC1, reducing H3K18ac at the TNF-α promoter (Bandyopadhaya et al. 2016). The findings with acetyl-CoA reduction, the primary substrate of histone acetylation, and the TNF-α transcription using UK5099 and ATP in 2-AA treated macrophages are in support of the bioenergetics disturbances observed in macrophages and their link to epigenetic modifications we have shown to be promoted by 2-AA (Bandyopadhaya et al. 2016)

(11) It is surprising that authors specifically target TNF alpha as a pro-inflammatory cytokine during tolerance. Various reports of cytokines and immune modulatory factors play a vital role in immune tolerance upon bacterial exposure. I would suggest authors perform cytokine profiling or check public data sets to specify their reason for choosing TNF alpha.

The choice of TNF-α is based on the results obtained in our previous study (Bandyopadhaya et al. 2016).

Bandyopadhaya, A., M. Kesarwani, Y. A. Que, J. He, K. Padfield, R. Tompkins, and L. G. Rahme. 2012. 'The quorum sensing volatile molecule 2-amino acetophenon modulates host immune responses in a manner that promotes life with unwanted guests', PLoS pathogens, 8: e1003024.

Bandyopadhaya, A., A. Tsurumi, D. Maura, K. L. Jeffrey, and L. G. Rahme. 2016. 'A quorum-sensing signal promotes host tolerance training through HDAC1-mediated epigenetic reprogramming', Nat Microbiol, 1: 16174.

Bandyopadhaya, A., A. Tsurumi, and L. G. Rahme. 2017. 'NF-kappaBp50 and HDAC1 Interaction Is Implicated in the Host Tolerance to Infection Mediated by the Bacterial Quorum Sensing Signal 2-Aminoacetophenone', Front Microbiol, 8: 1211.

Chaltel-Lima, L., F. Domínguez, L. Domínguez-Ramírez, and P. Cortes-Hernandez. 2023. 'The Role of the Estrogen-Related Receptor Alpha (ERRa) in Hypoxia and Its Implications for Cancer Metabolism', Int J Mol Sci, 24.

Lauterbach, M. A., J. E. Hanke, M. Serefidou, M. S. J. Mangan, C. C. Kolbe, T. Hess, M. Rothe, R. Kaiser, F. Hoss, J. Gehlen, G. Engels, M. Kreutzenbeck, S. V. Schmidt, A. Christ, A. Imhof, K. Hiller, and E. Latz. 2019. 'Toll-like Receptor Signaling Rewires Macrophage Metabolism and Promotes Histone Acetylation via ATP-Citrate Lyase', Immunity, 51: 997-1011 e7.

McCommis, K. S., and B. N. Finck. 2015. 'Mitochondrial pyruvate transport: a historical perspective and future research directions', Biochem J, 466: 443-54.

Tzika, A. A., C. Constantinou, A. Bandyopadhaya, N. Psychogios, S. Lee, M. Mindrinos, J. A. Martyn, R. G. Tompkins, and L. G. Rahme. 2013. 'A small volatile bacterial molecule triggers mitochondrial dysfunction in murine skeletal muscle', PloS one, 8: e74528.

Version 2

Abstract

Abstract

How bacterial pathogens exploit host metabolism to promote immune tolerance and persist in infected hosts remains elusive. To achieve this, we show that Pseudomonas aeruginosa (PA), a recalcitrant pathogen, utilizes the quorum sensing (QS) signal 2-aminoacetophenone (2-AA). Here, we unveil how 2-AA-driven immune tolerization causes distinct metabolic perturbations in macrophages’ mitochondrial respiration and bioenergetics. We present evidence indicating that these effects stem from decreased pyruvate transport into mitochondria. This reduction is attributed to decreased expression of the mitochondrial pyruvate carrier (MPC1), which is mediated by diminished expression and nuclear presence of its transcriptional regulator, estrogen-related nuclear receptor alpha (ERRα). Consequently, ERRα exhibits weakened binding to the MPC1 promoter. This outcome arises from the impaired interaction between ERRα and the peroxisome proliferator-activated receptor gamma coactivator 1-alpha (PGC-1α). Ultimately, this cascade results in diminished pyruvate influx into mitochondria and, consequently reduced ATP production in tolerized macrophages. Exogenously added ATP in infected macrophages restores the transcript levels of MPC1 and ERR-α and enhances cytokine production and intracellular bacterial clearance. Consistent with the in vitro findings, murine infection studies corroborate the 2-AA-mediated long-lasting decrease in ATP and acetyl-CoA and its association with PA persistence, further supporting this QS signaling molecule as the culprit of the host bioenergetic alterations and PA persistence. These findings unveil 2-AA as a modulator of cellular immunometabolism and reveal an unprecedented mechanism of host tolerance to infection involving the PGC-1α/ERRα axis in its influence on MPC1/OXPHOS-dependent energy production and PA clearance. These paradigmatic findings pave the way for developing treatments to bolster host resilience to pathogen-induced damage. Given that QS is a common characteristic of prokaryotes, it is likely that 2-AA-like molecules with similar functions may be present in other pathogens.

Introduction

Host tolerance is a fundamental mechanism of innate immunity. Studies have shown that innate immune cells following infection or exposure to microbial products may enter a state of immune tolerance characterized by reduced responsiveness towards microbial re-exposure a few hours later [1–3]. Epigenetic and signaling-based mechanisms have been proposed to be involved in host immune tolerance [1–3]. In recent years, several reports have highlighted the complex interplay between metabolic reprogramming and immunity [4, 5]. It was suggested that specific metabolic programs are activated in monocytes and macrophages upon exposure to infection and microbial products [6, 7]. However, in host tolerance, the molecular mechanisms underlying immunometabolic reprogramming mediated by bacterial pathogens remain poorly understood.

Metabolic pathways are essential for generating energy for various cellular functions, including those performed by immune cells [8, 9]. Macrophages use metabolic pathways to generate energy and metabolites to adapt to changing environments and stimuli, thereby enabling them to cope with the needs of a fluctuating immune response [7, 10]. Thus, properly functioning metabolic pathways are vital for immune cells to counteract pathogens. For instance, the crucial energy-carrying molecule adenosine triphosphate (ATP) is required to phagocytose pathogen-derived molecules efficiently.

Pseudomonas aeruginosa (PA), a recalcitrant ESKAPE (Enterococcus faecium, Staphylococcus aureus, Klebsiella pneumoniae, Acinetobacter baumannii, PA, and Enterobacter sp.) pathogen that causes acute and persistent infections, secretes virulence-associated low molecular weight signaling molecules, several of which are regulated by quorum sensing (QS) [11–14] and able to modulate host immune responses [15–18]. QS is a cell density-dependent signaling system bacteria use to synchronize their activities [11–14]. MvfR (a.k.a. PqsR), a critical QS transcription factor of PA, regulates the synthesis of many small molecules, including signaling molecules such as 2-aminoacetophenone (2-AA) [11, 19–21]. In vivo studies have demonstrated that 2-AA enables PA to persist in infected murine tissues by promoting innate immune tolerance through histone deacetylase 1 (HDAC1)-mediated epigenetic reprogramming [17, 18]. 2-AA tolerization reprograms the host inflammatory signaling cascade by maintaining chromatin in a “silent” state through increased HDAC1 expression and activity and decreased histone acetyltransferase (HAT) activity [18]. These changes also impact the protein-protein interaction between HAT and cyclic AMP response element-binding protein/HDAC1 and p50/p65 nuclear factor (NF)-κB subunits [18, 22]. The 2-AA-mediated tolerization permits PA to persist in infected tissues [17, 18], underscoring the difference from lipopolysaccharide (LPS)-mediated tolerization, which instead leads to bacterial clearance and involves different HDACs [23].

Mitochondria, the “powerhouse” of the cells, are crucial for the regulation, differentiation, and survival of macrophages and other immune cells [8]. Our group’s previous in vivo and in vitro studies indicated that 2-AA affects metabolic functions in skeletal muscle, which contains a high concentration of mitochondria [24–26]. Injection of 2-AA in murine skeletal muscle dampens the expression of genes associated with OXPHOS and the master regulator of mitochondrial biogenesis peroxisome proliferator-activated receptor-γ coactivator-1 beta (PGC-1β) [24]. Another PA QS molecule, 3-oxo-C12-HSL, attenuates the expression of PGC-1α [27]. PGC-1α and PGC-1β belong to the PPARγ family of inducible transcriptional coactivators and are known regulators of mitochondrial metabolism [28]. The coactivator PGC-1 proteins are involved in various cellular energy metabolic processes [29, 30], including but not limited to mitochondrial metabolism [31]. Recent studies have also emphasized the role of estrogen-related nuclear receptor α (ERRα) in coordinating metabolic capacity with energy demand in health and disease [32]. PGC-1α activates ERRα transcriptional activity through protein-protein interaction, which enhances the expression of ERRα [33, 34] and other mitochondrial genes involved in lipid metabolism and OXPHOS [35]. PGC-1α/ERRα axis has been extensively studied in cancer and established as a central regulatory node of energy metabolism that induces the global expression of genes involved in mitochondrial biogenesis and functions [36–38]. ERRα has been shown to occupy the promoter regions of many genes participating in the TCA cycle and OXPHOS [39, 40], including the mitochondrial pyruvate carrier (MPC1). In human renal carcinoma cells, PGC-1α/ERRα interaction results in efficient activation of MPC1 expression and transport of pyruvate into mitochondrion for efficient OXPHOS and energy production [41].

Given that our previous studies pointed to the 2-AA effect on mitochondrial functions [24–26] and that mitochondria are crucial for the regulation, differentiation, and survival of macrophages and other immune cells [8], we investigated how 2-AA may mediate cellular metabolic reprogramming in tolerized immune cells by examining the link between the PGC-1α/ERRα axis and 2-AA-mediated immune tolerance. Our findings uncovered the crucial action of 2-AA on energy homeostasis and metabolism in tolerized immune cells through perturbances of the ERRα/PGC-1α interaction, MPC1-mediated pyruvate transport and the production of the key energy metabolism molecules, ATP and acetyl-CoA and their association with the persistence of PA in mammalian tissues.

Results

Tolerization by 2-AA impacts the generation of crucial energy metabolites in macrophages

Following infection or exposure to microbial products, innate immune cells may enter a state of tolerance several hours later, as depicted by their diminished responsiveness or unresponsiveness to microbial re-exposure [1–3, 42]. Given that 2-AA tolerized cells are non-responsive to 2-AA re-exposure, we investigated the impact of 2-AA on energy homeostasis and metabolism in non-tolerized and tolerized BMDM cells (Fig. 1). We quantified the levels of ATP and acetyl-CoA, which is a fuel for ATP production, in a series of in vitro experiments in which murine BMDM cells were exposed to 2-AA (1st 2-AA exposure) or re-exposed (2nd 2-AA exposure) to respectively determine their activation, and tolerance as well as memory to this QS molecule (Fig. 1A). Initially, following the 1st 2-AA exposure (black bars), a significant increase in the concentrations of intracellular ATP (Fig. 1B) and acetyl-CoA (Fig. 1C) was observed in BMDM cells exposed to 2-AA for 1 and 6 h compared to their corresponding naïve control cells (grey bars). However, cells exposed to 2-AA for 48 h (black bar) appeared to have entered a state of tolerance as ATP and acetyl-CoA levels were decreased compared to 2-AA exposures for a shorter time (Figs. 1B and 1C). The tolerance of these cells as well as their memory to 2-AA was determined by their responsiveness to a 2nd 2-AA exposure (Fig. 1A). To achieve this, these cells were washed to remove 2-AA and allowed to rest for 24 or 106 h in the absence of 2-AA before receiving the 2nd 2-AA exposure for 1 or 6 h (Fig. 1A). Figures 1B and 1C (red bars) show the unresponsiveness of BMDM cells to 2-AA re-exposure as evidenced by the similar levels of ATP and acetyl-CoA compared to cells exposed to 2-AA for 48 h (1st exposure black bars). The sustained unresponsiveness of the cells to re-exposure indicates that the cells exhibited innate immunological memory, which could not be reverted by 2-AA re-exposure. Similar responsiveness patterns of ATP and acetyl-CoA as with BMDM cells were observed in murine macrophage RAW 264.7 cells (Figs. S1A and S1B) or human monocyte THP-1 cells (Figs. S1C and S1D) following exposure or re-exposure to 2-AA. These findings further support our observations with BMDM cells and confirm that either cell type can be used to study responses to 2-AA. These results reinforce the notion that 2-AA impacts crucial energy metabolites in macrophages.

2-AA tolerization decreases crucial metabolites of cellular energy and affects mitochondrial respiration in mouse BMDM.

(A) Schematic representation showing experimental design: naïve cells were exposed to 2-AA for 1, 6, or 48 h (black). Cells exposed to 2-AA (400 µM) for 48 h were washed, rested for 24 or 106 h, and re-exposed (200 µM) for 1 or 6 h (red), respectively. The same color code, black cells after 1st exposure and red cells after 2nd exposure, was kept throughout the manuscript with corresponding controls in grey and pink, respectively. The levels of (B) ATP and (C) acetyl-CoA in BMDM cells after 1st and 2nd 2-AA exposure. (D) Real-time oxygen consumption rate (OCR) traces were recorded using a Seahorse XF analyzer and normalized to protein content. Cells were exposed to 2-AA for 1 or 24 h (black), washed and rested for 24 h and re-exposed for 1 h (red). Mitochondrial respiratory parameters, (E) basal respiration, and (F) maximal respiration. Data are presented as mean ± SD, n ≥ 4, *p < 0.05, **p < 0.01, ***p < 0.001, and ns indicates no significant difference. One-way ANOVA followed by Tukey’s post hoc test was applied.

Moreover, we sought to determine whether 2-AA-tolerized BMDM cells would respond to heterologous stimulus. To test this, we re-exposed the cells to lipopolysaccharide (LPS), an outer membrane bacterial component that strongly induces macrophages. LPS re-exposure of the 2-AA tolerized BMDM cells did not significantly (p = 0.9) increase ATP levels (Fig. S2), indicating that 2-AA cross-tolerized the cells to this heterologous bacterial immunostimulant. On the other hand, as expected, we observed increased ATP levels following LPS stimulation of the non-tolerized BMDM cells (Fig. S2). The broad spectrum of 2-AA cross-tolerization strongly suggests an interplay of epigenetic mechanism rather than ligand-receptor mediated signaling-based mechanism in bringing out the observed 2-AA-mediated tolerance.

Tolerization by 2-AA leads to a quiescent state by reducing cell bioenergetics and compromising OXPHOS

To further investigate the 2-AA’s impact on the bioenergetics of macrophages, we assessed by Seahorse assay the oxygen consumption rate (OCR) as an index of OXPHOS in 2-AA exposed (for 1 and 24 h) and re-exposed (2nd exposure for 1 h) BMDM cells (Fig. 1D). Values of basal (Fig. 1E) and maximal (Fig. 1F) mitochondrial respiration and spare respiratory capacity (Fig. S3) interpreted from these measurements revealed that 1 h 2-AA exposure of naïve cells significantly increased OCR compared to naïve control cells. At 24 h post-2-AA exposure, cells had a significantly reduced basal OCR level than the control BMDM cells, indicating decreased OXPHOS and an overall quiescent phenotype of tolerized macrophages. Re-exposure of tolerized macrophages with 2-AA also did not augment maximal and basal OCR (Figs. 1D-1F), indicating the unresponsiveness of tolerized BMDM cells. Ultramicroscopic examination of RAW 264.7 cells exposed to 2-AA for 48 h although shows the same number of mitochondria as in control cells, their mitochondrial morphology is altered, appearing smaller and round and having reduced cristae indicating dysfunctional mitochondria (Fig. S4).

These OCR data are consistent with previous direct measurements of ATP (Fig. 1B) and acetyl-CoA (Fig. 1C) and support the notion that 2-AA tolerization induces a quiescent phenotype in these cells characterized by defective OXPHOS. These findings confirm that 2-AA mediates energy homeostatic and metabolic alterations in the immune cells and that 2-AA tolerization arrests these cells in a sustained 2-AA-unresponsive state.

The pyruvate transport into mitochondria is decreased in tolerized macrophages

Since acetyl-CoA and ATP levels were reduced due to 2-AA tolerization, we investigated whether the mitochondrial dysfunction observed is related to pyruvate metabolism. Pyruvate links glycolysis with mitochondrial production of acetyl-CoA and ATP via the tricarboxylic acid (TCA) cycle and OXPHOS [7]. RAW 264.7 cells initially exposed to 2-AA for 1 or 3 h (black bars) exhibited significantly higher concentrations of cytosolic and mitochondrial pyruvate compared to the corresponding naïve control cells (Fig. 2A). Pyruvate significantly decreased over time in cytosolic and mitochondrial fractions as cells were entering into the tolerized state (Fig. 2A). Tolerized cells remain unresponsive to 2-AA re-exposure (red bars), exhibiting no significant change in the concentration of pyruvate in the cytosolic or mitochondrial fraction (Fig. 2A). A similar result was observed in THP-1 cells (Figs. S1E and S1F).

2-AA perturbs the mitochondrial MPC1-mediated import and metabolism of pyruvate.

(A) Cytosolic and mitochondrial pyruvate levels following 2-AA exposure (black) or re-exposure (red) and corresponding controls in grey and pink, respectively, for indicated time points. (B) Representative Western blot and results of densitometric analysis of MPC1 protein levels following 2-AA exposure or re-exposure for indicated time points. β-actin was used as a control. Corresponding controls are shown in grey or pink, respectively. Mean ± SD is shown, n = 3, ***p < 0.001, ****p < 0.0001, and ns indicates no significant difference. One-way ANOVA followed by Tukey’s post hoc test was applied.

To gain a better understanding of the 2-AA-mediated reductions in mitochondrial pyruvate, ATP production, and acetyl-CoA levels, we sought to determine the involvement of the MPC1 [43] that is crucial in transporting pyruvate from the cytosol into the mitochondria [44]. Western blot studies of whole cell lysates of RAW 246.7 cells exposed to 2-AA showed a significant decrease over time in the abundance of MPC1 protein, and tolerized cells remain unresponsive to 2-AA re-exposure (Fig. 2B). Together, these findings suggest that 2-AA tolerization dysregulates MPC1-mediated transport of the pyruvate into mitochondria.

2-AA tolerization suppresses ERRα binding to the MPC1 promoter and the interaction between PGC-1α and ERRα

Given that the PGC-1α/ERRα axis is the transcriptional regulatory node that activates the expression of MPC1 [33], we determined the protein levels of ERRα. Western blot studies of whole cell lysates showed a significant decrease over time in the abundance of ERRα protein following 1st exposure of RAW 246.7 cells to 2-AA, while tolerized cells remain unresponsive to 2-AA re-exposure (Fig. 3A). By assessing the cytosolic and nuclear abundance of ERRα in tolerized cells we found that both fractions had lower levels of ERRα protein than their corresponding control cells (Fig. 3B).

2-AA-mediated macrophage tolerization deranges PGC-1α/ERRα-dependent metabolic programing.

(A) Representative Western blot and results of densitometric analysis of ERRα protein levels following 2-AA exposure or re-exposure for indicated time points. β-actin was used as a control. (B) Western blots and its corresponding densitometric analysis of ERRα in cytoplasmic or nuclear lysates of tolerized macrophages exposed to 2-AA for 24 h or not exposed to 2-AA. (C) ChIP-qPCR assay of ERRα binding at the MPC1 promoter in RAW 264.7 tolerized macrophages exposed to 2-AA (200 μM) for 24 h (black) compared to untreated control macrophages (grey). IgG served as a negative control (D) Representative Western blot of co-immunoprecipitation (co-IP) studies of ERRα and PGC-1α in nuclear extracts of 2-AA tolerized (24 h) and control RAW 264.7 cells. Pull down with IgG served as a negative control. 2-AA tolerized macrophages are shown in black, and untreated control macrophages in grey. Mean ± SD is shown, n ≥ 3, *p < 0.05, **p < 0.01, ****p < 0.0001, and ns indicates no significant difference. One-way ANOVA followed by Tukey’s post hoc test was applied.

Furthermore, because an ERRα binding putative DNA motif (TNAAGGTCA) has been identified upstream of the MPC1 promoter in humans [43], and we found the same motif to be present 1.5 kB upstream of the transcription start site of mouse MPC1 promoter, we analyzed if tolerization led to reduced ERRα binding to the MPC1 promoter motif TNAAGGTCA. Chromatin immunoprecipitation (ChIP) assay followed by qPCR analysis revealed that in tolerized cells, ERRα binds approximately 4-fold less efficiently to the putative binding site on the MPC1 than in corresponding control cells (Fig. 3C).

Since the transcriptional activity of ERRα is regulated through a protein-protein interaction with the transcriptional coactivator PGC-1α [33], we assessed whether the formation of the PGC-1α/ERRα complex is affected in 2-AA tolerized cells. Using paraformaldehyde fixed nuclear lysates of RAW 264.7 tolerized and naïve cells, we performed a co-immunoprecipitation (co-IP) of PGC-1α and reversed co-IP with ERRα followed by immunoblotting (Fig. 3D). As shown in Figure 3D less ERRα was detected when PGC-1α was pulled down in presence of 2-AA than in naïve control cells. Moreover, reverse co-IP with ERRα protein showed lower levels of PGC-1α in the presence of 2-AA than in naïve control cells (Fig. 3D). Control IgG showed no detectable levels of ERRα following nuclear fraction pull down. These results indicate that in the tolerized macrophages, PGC-1α/ERRα interaction is impaired (Fig. 3D).

Moreover, since PGC-1α enhances the transcriptional activity of ERRα through their interaction, and ERRα transcription is regulated via an autoregulatory loop [34], we examined the effect of tolerization on the transcription of PGC-1α, ERRα, and its target gene MPC1. Using RAW 264.7 cells infected with the wild type PA strain PA14 or isogenic ΔmvfR mutant, which does not produce 2-AA [21] we observed lower RNA transcript levels of MPC1 (Fig. 4A), ERRα (Fig. 4B) and PGC-1α (4C) in PA14 compared to ΔmvfR. The addition of 2-AA to ΔmvfR decreased the levels of these gene transcripts to levels similar to the PA14 infection condition, supporting the role of 2-AA in effect. To test if the observed effects are due to MPC1-dependent reduction of mitochondrial ATP generation, we supplemented the macrophages with ATP. The findings indicate that the addition of ATP in PA14-infected cells elevated the transcript levels of ERRα, MPC1, and PGC-1α, reaching levels similar to those observed in ΔmvfR infected macrophages (Fig. 4A, 4B, and 4C). These results indicate that reduced MPC1 function is due to 2-AA tolerization on the transcriptional activation of MPC1 through the PGC-1α/ERRα axis.

Increased intracellular burden in macrophages is associated with decreased expression of MPC1, ERR-α, and TNF-α genes.

Real-time PCR analysis of MPC1 (A), ERR-α (B), and PGC-1α (C) expression in RAW 246.7 macrophages infected with PA14 or ΔmvfR in the presence or absence of exogenous addition of 2-AA or ATP for 6 h as indicated. Transcript levels were normalized to 18srRNA. PA14-infected cells served as controls. (D) The intracellular burden of PA14 or ΔmvfR of infected macrophages in the presence or absence of exogenous addition of 2-AA, UK5099, or ATP. Untreated cells infected with PA14 were set as 100%. (E) Real-time PCR analysis of TNF-α expression in RAW 246.7 macrophages infected with PA14 or ΔmvfR in the presence or absence of exogenous addition of 2-AA, ATP, or UK5099. Transcript levels were normalized to 18srRNA. PA14-infected cells served as controls. The compound concentration used for UK5099 was 10 µM and ATP 20 µM. Mean ± SD is shown, n = 3, * p < 0.05, **p < 0.01, ***p < 0.001, ****p < 0.0001***, and ns indicates no significant difference. One-way ANOVA followed by Tukey’s post hoc test was applied.

2-AA tolerization impairs macrophage-mediated intracellular bacterial clearance through decrease in MPC1-mediated pyruvate import, ATP, and TNF-α levels

2-AA-mediates persistence of PA in vivo, dampens proinflammatory responses, and increases the intracellular burden of this pathogen in macrophages via epigenetic modifications [17, 18, 26]. Here, we used RAW 264.7 cells infected with the PA14 or the 2-AA deficient ΔmvfR mutant to assess the clearance of PA by macrophages. Macrophages infected with PA14 showed increased bacterial burden than the cells infected with ΔmvfR, and 2-AA addition to ΔmvfR led to increased bacterial burden (Fig. 4D). Furthermore, we used UK5099 and ATP to interrogate whether MPC1 mediated mitochondrial pyruvate import and bioenergetics are linked to the clearance of PA intracellular burden. The addition of the UK5099 inhibitor strongly enhanced the bacterial intracellular burden in ΔmvfR infected macrophages compared to the non-inhibited ΔmvfR infected cells, reaching a similar burden to those infected with PA14 (Fig. 4D). Conversely, exogenously added ATP to macrophages infected with PA14 strongly reduced the PA intracellular burden (Fig. 4D).

Given that 2-AA tolerization decreases the expression of pro-inflammatory cytokine TNF-α by hypo-acetylating the core histone 3 lysine 18 acetylation (H3K18ac) mark at TNF-α promoter [18], we investigated the link between bioenergetics and TNF-α expression in infected macrophages. UK5099, or ATP was added exogenously to infected macrophages with PA14 or ΔmvfR (Fig. 4E). As shown in Figure 4E, PA14 infected cells showed lower TNF-α transcript levels compared to the ΔmvfR infected cells. Supplementation of 2-AA to ΔmvfR infected cells led to a decrease in the TNF-α transcript levels (Fig. 4E). The addition of the MPC1 inhibitor, UK5099 in ΔmvfR infected cells counteracted the increase in TNF-α transcript levels observed in ΔmvfR infected cells in the absence of UK5099 (Fig. 4E). Conversely, exogeneous addition of ATP to PA14 infected cells enhanced the transcription of TNF-α compared to the untreated PA14 infected cells, while no difference in TNF-α expression levels were observed in ΔmvfR infected macrophages in presence or absence of ATP (Fig. 4E).These findings strongly suggest that 2-AA tolerization severely alters macrophages’ ability to facilitate the clearance of PA intracellular burden, via the reduction in MPC1-mediated pyruvate import into mitochondria, ATP levels and TNF-α transcription.

In vivo studies corroborate the 2-AA-mediated decrease in vitro of the central metabolic fuel acetyl-CoA and the energy-carrying molecule ATP and their association with PA persistence

To determine whether the decrease in the key metabolites mediated by 2-AA is also observed in vivo during infection, we quantified the levels of ATP and acetyl-CoA in murine spleen tissues at 1, 5, and 10 days (Fig. 5). This organ was selected for our immunometabolic studies due to its role in regulating not only local but also systemic (whole-body) immune responses, facilitated by various immune cells including macrophages [45]. Mice were infected with PA strain PA14, or isogenic mutant ΔmvfR, which does not produce 2-AA [21]. It is important to note that it is not possible to generate or utilize a bacterial mutant that is defective in 2-AA only because 2-AA is formed by spontaneous decarboxylation rather than by an enzyme-catalyzed reaction [46–48]. Therefore, animals that were either infected with ΔmvfR and received 2-AA (ΔmvfR+2-AA) at the time of infection or uninfected mice injected with a single dose of 2-AA served as direct controls (Fig. 5). Naïve and sham mice groups served as additional basal controls (Fig. 5).

2-AA promotes a long-lasting decrease in ATP, acetyl-CoA levels, and bacterial persistence in PA-infected mice.

(A) ATP and (B) acetyl-CoA concentrations in the spleens of mice infected with PA wild type (PA14), the isogenic mutant ΔmvfR, ΔmvfR injected with 2-AA at the time of infection (ΔmvfR + 2-AA), or non-infected but injected with 2-AA (6.75 mg/kg), (C) Bacterial burden in muscles expressed as CFU count was analyzed using the Kruskal–Wallis non-parametric test with Dunn’s post-test; ***p < 0.001, and ns indicates no significant difference. Control mice groups: naïve were not given 2-AA; mice receiving 2-AA were given a single intraperitoneal injection of 2-AA; sham represents a burn/PBS group since the burn and infection model was used. Results of three independent replicates with 4 mice per group for 1, 5, and 10 d are shown. Means ± SD are shown, *p < 0.05, **p < 0.01, ***p < 0.001, and ns indicate no significant difference. One-way ANOVA followed by Tukey’s post hoc test was applied.

Infection with PA14 or uninfected mice injected with 2-AA led to a significant decrease in both ATP (Fig. 5A) and acetyl-CoA (Fig. 5B) concentrations in spleen tissues compared to ΔmvfR infected mice that sustained higher ATP and acetyl-CoA levels similar to naïve and sham control groups across the time points tested. However, in mice infected with ΔmvfR and 2-AA injected (ΔmvfR+2-AA) at the time of infection, ATP and acetyl-CoA concentrations in the spleen decreased to levels comparable to those of PA14-infected animals, strongly indicating the biological function of 2-AA in decreasing these key metabolites. These in vivo findings further support the adverse action of 2-AA on host energy homeostasis and metabolism observed in in vitro studies.

The metabolic alterations observed are associated with the host tolerance to PA persistence (Fig. 5C). Using Pseudomonas isolation agar plates, we evaluated the bacterial load at the infection site over the course of 10 days, obtaining samples at 1, 5, and 10 days. At one day post-infection, mice infected with PA14, ΔmvfR, or ΔmvfR+2-AA exhibited no difference in bacterial burden (Fig. 5C), verifying the ability of all strains to establish infection. At 5- and 10 days post-infection, mice infected with PA14 or ΔmvfR+2-AA sustained the bacterial burden at a significantly higher bacterial burden over time compared to those infected with ΔmvfR (Fig. 5C).

To strengthen the relevance of our in vivo data, we performed additional in vivo experiments. In this set of in vivo studies, mice received the first exposure to 2-AA by injecting 2-AA only and the 2nd exposure through infection with PA14 or ΔmvfR four days post-2-AA injection. As shown in the supplementary Figure S5, the levels of ATP and acetyl CoA in the spleen of infected animals and the enumeration of the bacterial counts were similar between PA14 or ΔmvfR receiving the 1st 2-AA exposure and agree with the “one-shot infection” findings presented in Figure 5 with the PA14 or ΔmvfR +2-AA infected mice or those receiving 2-AA only. These results are consistent with our previous findings, showing that 2-AA impedes the clearance of PA14 [17, 18] and provide compelling evidence that the metabolic alterations identified may favor PA persistence in infected tissues.

Discussion

The PA signaling molecule 2-AA that is abundantly produced and secreted in human tissues [21, 22] is the first QS molecule that epigenetically reprograms immune functions, promotes immune tolerance, and sustains PA’s presence in host tissues [17, 18, 22, 26]. The present study provides insights into the mechanistic actions of 2-AA on cellular metabolism that contribute to host immune tolerance to PA persistence. We uncover that this signaling molecule causes distinct metabolic alterations in macrophages’ mitochondrial respiration and energy production promoted via the PGC-1α/ERRα axis and MPC1-dependent OXPHOS that links the TCA cycle to the production of ATP and energy homeostasis (Fig. 6). Our results show that 2-AA tolerization decreases ATP the crucial energy metabolite and the histone acetylation metabolite acetyl-CoA and in vivo, implicating the importance of the key energy-producing mitochondrial process in immune tolerance. Although macrophages respond to the first exposure to 2-AA by enhancing ATP and acetyl-CoA production, long-term exposure to 2-AA leads to a tolerized state characterized by sustained reduced ATP and acetyl-CoA concentrations, a quiescent bioenergetic state with reduced mitochondrial pyruvate uptake and unresponsiveness to a 2nd exposure.

Proposed mechanism by which 2-AA impairs bioenergetics through the inhibition of MPC1-mediated pyruvate transport into mitochondria and its impact on the PGC-1α/ERRα axis.

2-AA tolerized macrophages exhibit diminished pyruvate levels in mitochondria due to the decreased expression of MPC1, a consequence of the 2-AA impact on the interaction of ERRα with the transcriptional coactivator PGC-1α for the effective transcription of ERRα since ERRα controls its own transcription and that of MPC1. In the presence of 2-AA, the weakened interaction between ERRα and PGC-1α results in reduced expression of MPC1 and ERRα. The reduction in mitochondrial pyruvate levels leads to decreased acetyl-CoA and ATP levels, which modulate HDAC1- and HAT-catalyzed remodeling of H3K18 acetylation. The diminished levels of this epigenetic mark have previously been associated with an increased intracellular presence of bacteria in macrophages, as demonstrated by our group [18]. Pathways, proteins, and metabolites that are negatively affected are indicated in red, while positively affected are denoted in black. The figure was created using BioRender.com.

Our results reveal that the decrease of ATP and acetyl-CoA in tolerized macrophages results from the 2-AA-mediated perturbation of the physical interaction between ERRα and PGC-1α. The PGC-1α/ERRα axis has not previously been implicated in tolerization but has been extensively studied in cancer, underscoring the novelty of our findings. Studies in cancer cells have shown that the physical interaction of human ERRα with PGC-1α strongly enhances the binding of ERRα to DNA to induce the activation of the targeted genes [33], including MPC1, responsible for the import of cytosolic pyruvate into mitochondria and PDH, that catalyzes the conversion of pyruvate to acetyl-CoA in mitochondria. Inhibition of ERRα has also been shown in cancer cell metabolism studies to interfere with pyruvate entry in mitochondria by inhibiting the expression of MPC1 [49], and inhibition of ERRα led to decreased expression of PGC-1α-controlled expression of mitochondrial genes in mice brains [45]. Indeed, we find that the transcription of ERRα, PGC-1α, and MPC1 genes and pyruvate uptake into mitochondria is reduced in tolerized macrophages. Given that pyruvate is a primary carbon source in the TCA cycle that fuels OXPHOS and the production of ATP into mitochondria, it explains the reduced ATP and acetyl-CoA levels observed in tolerized macrophages. Future studies will focus on unveiling the pathways by which macrophages respond to the bioenergetic changes identified and decipher their effect on the pyruvate cycle in host tolerance to infection [50–52].

We show that the unprecedented action of a QS bacterial small signaling molecule on the interaction between ERRα and PGC-1α in the nucleus is hampered. Although it remains to be elucidated whether 2-AA interferes directly or indirectly with the PGC-1α/ERRα axis, our data bring up the possibility that the reduced production of acetyl-CoA may be responsible since cholesterol is a known coactivator/ligand of ERRα [53], is generated from acetyl-CoA [53]. This possibility may also explain the reduced expression of ERRα and lower abundance since cholesterol is also required for its activation and autoregulation. However, it is also possible that 2-AA may antagonize the interaction between ERRα and PGC-1α by binding to ERRα. Small molecules as antagonists of ERRα have been reported previously [54]. The weaker binding of ERRα to the MPC1 promoter site, and consequently reduced MPC1 expression, explains the reduced pyruvate presence in mitochondria and that of ATP and mitochondrial quiescence. These findings open avenues for exploring novel therapeutic strategies through the overexpression of ERRα to counteract the tolerization induced by 2-AA and enhance clearance of PA infection. Interestingly, ERRα overexpression on breast cancer metastases promotes an efficient antitumor immune response selectively in the bone [55].

The observed cellular metabolic perturbances also align with our previous studies in skeletal muscle, which pointed to mitochondrial dysfunction triggered by 2-AA [24, 25]. Interestingly, the PA QS LasR-regulated homoserine lactone molecule, 3-oxo-C12-HSL, was reported to attenuate the expression of PGC-1α and decrease the mitochondrial respiratory capacity in lung epithelial cells [27]. Moreover, another PA QS molecule, PQS, also regulated by LasR, was recently shown to induce organelle stress, including mitochondria, by disrupting the mitochondrial membrane potential in human macrophages [56]. As opposed to 2-AA, which dampens the pro-inflammatory response, PQS increases proinflammatory cytokines [56], consistent with the fact that PQS is an acute infection type molecule unrelated to chronic/persistent infections.

Previously, we reported that 2-AA tolerization induces histone deacetylation via HDAC1, reducing H3K18ac at the TNF-α promoter [18]. The findings with acetyl-CoA reduction, the primary substrate of histone acetylation, and the TNF-α transcription using UK5099 and ATP in 2-AA treated macrophages are in support of the bioenergetics disturbances observed in macrophages and their link to epigenetic modifications we have shown to be promoted by 2-AA [18]. Macrophages exposed to 2-AA in the presence of exogenous ATP showed improved intracellular bacterial clearance and enhanced TNF-α levels, supporting the epigenetic interconnection of macrophages and cellular ATP responsiveness levels against infection. The exogenous addition of UK5099 that reverted the efficiency of the mvfR infected macrophages to clear the PA intracellular burden and the counteraction to the 2-AA effect by ATP addition that increases the clearance of PA intracellular burden support the importance of this energy-carrying molecule in PA persistence.

The results with the MPC1 inhibitor, UK5099, suggest that the availability of pyruvate could underlie the mechanism of 2-AA-regulated HDAC/HAT-dependent control of transcription of pro-inflammatory mediators and bacterial clearance [57]. These results align with our previous findings, showing that establishing bacterial intracellular burden in macrophages is HDAC1 dependent [57]. It remains to be elucidated if HDAC/HAT-mediated histone modification directly regulates the expression of OXPHOS genes in 2-AA tolerized macrophages, as it was shown that histone deacetylation downregulates OXPHOS in persistent M. tuberculosis infection [58, 59]. Given the complexity of 2-AA-mediated long-term effects, future omics studies combined with immune profiling may aid in deciphering the possible network of co-factors across different subpopulations of immune cells and the immunometabolic reprogramming related to 2-AA.

We have shown that although 2-AA tolerization leads to a remarkable increase in the survival rate of infected mice, it permits an HDAC1-dependent sustained presence of PA in mice tissues and intracellularly in macrophages [17, 18, 22, 26]. Here, we use 2-AA-producing and isogenic non-producing PA strains to confirm the 2-AA-mediated decrease in the central metabolite acetyl-CoA and the energy-carrying molecule ATP during infection. These murine infection studies provide strong evidence of the 2-AA biological relevance in reducing these metabolites in vivo. Taking together our previous and current in vivo findings provide compelling evidence that the metabolic alterations identified favor PA persistence in infected tissues.

That 2-AA permits PA to persist in infected tissues despite rescuing the survival of infected mice underscores its difference from that of LPS tolerization, which results in bacterial clearance [23] via the mechanism that relies on a set of different HDAC enzymes. LPS tolerization predominantly involves changes in H3K27 acetylation [60], while 2-AA tolerization involves H3K18 modifications [22]. The 2-AA-mediated effects reported here are also distinct from the epigenetic-metabolic reprogramming mediated by LPS [61], which promotes endotoxin tolerance in immune cells by upregulating glycolysis and suppressing OXPHOS [62]. Although 2-AA and LPS implicate different components and lead to different outcomes, both involve epigenetic mechanisms and immune memory.

Metabolic reprogramming upon infection may be pathogen-specific, with each pathogen impacting specific metabolic pathways that better fit its respective metabolic needs [7]. This was shown for infections of various tissues with Chlamydia pneumonia, Legionella pneumophila, Mycobacterium tuberculosis, and Salmonella [63–66] and, by our group, for PA infections [17, 18, 24]. It would be interesting, however, to test whether 2-AA affects the killing efficacy of macrophages against other pathogens, as emerging evidence shows that synergistic or antagonistic interactions between clinically relevant microorganisms and host have important implications for polymicrobial infectious diseases [67].

While in this study, we focused on the role of ERRα mainly in pyruvate metabolism; future studies are needed to reveal other ERRα/PGC-1α axis-dependent metabolic and immune pathways are modulated by 2-AA. In this view, the cholesterol synthesis pathway that relies on acetyl-CoA precursor, as cholesterol is a known coactivator/ligand of ERRα [53] and fatty acid catabolism, generating acetyl-CoA, and PGC-1α regulates fatty acid oxidation [68].

This study unveils the unprecedented actions of a QS bacterial molecule in orchestrating cellular metabolic reprogramming in addition to the epigenetic reprogramming reported previously and has also been shown to promote a long-lasting presence of PA in the host [17, 18, 26]. That 2-AA cross-tolerized macrophages to LPS corroborates our previous findings on the implication of epigenetic mechanism [18] rather than ligand-receptor mediated signaling-based mechanism and raises the possibility that this QS molecule may confer non-specific cross-protection. This is an aspect we plan to investigate in the future.

The reported immunometabolic reprogramming contributes to a better understanding of the molecular and cellular mechanisms, biomarkers, and functional significance that may be involved in immune tolerance to persistent infection, providing for designing and developing innovative therapeutics and interventions. These approaches can focus on promoting host resilience against bacterial burden and safeguarding patients from recalcitrant persistent infections.

Materials & Methods

Ethics statement

The animal protocol was approved by the Institutional Animal-Care and Use Committee (IACUC) of Massachusetts General Hospital (protocol no. 2006N000093). No randomization or exclusion of data points was applied.

Conditions used for the exposures of the cells to 2-AA

Mouse BMDM cells were isolated from the femur of CD1 6 weeks old male and female mice and used to estimate the levels of various metabolites (Fig. 2A) following exposure and re-exposure to 2-AA (Fig. 1A). The 1st exposure of naïve BMDMs (106/mL in 6-well plates) was achieved with 400 µM 2-AA for 1, 6, or 48 h. Cells exposed for 48 h were used for the 2nd 2-AA exposure. These cells were washed to remove residual 2-AA, left in the medium in the absence of 2-AA for either 24 or 106 h, and re-exposed to 2-AA (200 µM) for 1 or 6 h, respectively. For RAW 264.7 (106/mL in 6-well plates) and THP-1 cells (106/mL in 6-well plates), the 2-AA concentration and conditions used were the same as those used for BMDM cells. The hours of the 1st round and 2nd round of exposure to 2-AA are indicated in Figure S1.

For the LPS stimulation studies, cells were stimulated with 400 µM 2-AA or 100 ng/mL LPS for 6 h and 800 µM of 2-AA for 48 h. Cells receiving 1st exposure for 48 h were washed and re-exposed to 400 µM of 2-AA or 100 ng/mL LPS for 6 h.

ATP and acetyl-CoA quantifications

The levels of ATP acetyl-CoA, and were assessed in BMDM and RAW 264.7 macrophage cells following exposure to 2-AA at the times indicated and by utilizing the ATP Assay kit (cholorimetric/fluorometric) (#ab83355, Abcam) and the PicoProbe AcCoA assay kit (ab87546, Abcam), respectively, according to the manufacturer’s instruction. Quantifications of ATP and acetyl-CoA were performed in triplicates.

ATP

For ATP determination, macrophages were lysed and subsequently centrifuged at 14,000 rpm for 10 min at 4 °C. The supernatants were transferred to a fresh Eppendorf tube. Standards and cell supernatants of 50 µL were added to a 96-well plate suitable for fluorescent analysis (black sides, clear bottom). A reaction mixture containing an ATP converter, probe, buffer, and developer mix was then added to all wells (50 µL) and incubated away from light at room temperature for 30 min. ATP quantification was conducted fluorometrically at 535/587 nm using a plate reader and the following settings: λex = 535 nm; λem = 587 nm. Fluorescence was measured using a microplate reader (Tecan Group Ltd, Männedorf, Switzerland)

Acetyl-CoA

Acetyl-CoA content was assessed by first deproteinizing total cell fractions of macrophages using the perchloric acid and then centrifuging at 14,000 rpm for 10 min at 4 °C. 50 µL of cells’ supernatant sample CoASH were quenched to correct the background generated by free CoASH. Following the homogenization procedure described above, the samples were diluted with the reaction mix, and fluorescence was quantified using a plate reader and the settings as above with ATP.

Seahorse Assays

Seahorse analysis was performed according to previously published protocols [69]. Briefly, freshly prepared BMDM cells were reseeded in complete RPMI-1640 cells using Seahorse plates (Agilent cat. no. 103729100) at a density of 5 × 104 cells per well. Cells were exposed to 400 µM 2-AA for 1 h or 24 hours (1st exposure). Cells receiving 1st exposure for 24 h were washed and re-exposed to 200 µM of 2-AA for 1 h. Naïve cells were used as control. Prior to initiating Seahorse measurements, cells were washed, and Seahorse XF DMEM supplemented with 2 mM glutamine (Gibco cat. No. 25030-081) was added to each well. Cells were allowed to stabilize in a 37°C incubator without CO2 for 1 h. The Seahorse cartridge was hydrated and calibrated as per the manufacturer’s instructions. The Mitochondrial stress test kit (Agilent cat. no. 10395-100) was used (oligomycin at 1 μM, FCCP (1.5 µM), rotenone (0.5 μM), and antimycin A at (0.5 μM)) with slight modifications according to the published protocol [69] that included injection of 25 mM glucose, and sodium pyruvate (1 μM). All samples N= 4 were run in a Seahorse XFe96 Analyzer, and data were analyzed using Wave and plotted using GraphPad Prism software.

Isolation of cytosolic and mitochondrial fraction and pyruvate quantification

Pyruvate is produced in the cytosol and is transported into the mitochondria. RAW 264.7 cells were plated at 1 × 106 to incubate overnight at 37°C in a CO2 incubator. Cells were exposed to 2-AA for 1, 3, 24, or 48 h or re-exposed for 1 h as described above. Mitochondria and cytosolic fractions of each group were isolated utilizing the Mitochondria Isolation Kit for cultured cells (#ab110170, Abcam) following the manufacturer’s protocol. Briefly, cells were collected with a cell lifter and pelleted by centrifugation at 1,000 g, frozen, and then thawed to weaken the cell membranes. The cells were resuspended in Reagent A and transferred into a pre-cooled Dounce Homogenizer. The homogenates were centrifuged at 1,000 g for 10 min at 4°C and saved as supernatant #1 for the cytosolic fractions. The pellet was resuspended in Reagent B, followed by repeat rupturing and centrifugation. The pellet was collected and resuspended in 500 μL of Reagent C supplemented with Protease Inhibitor cocktails (P8340, Sigma Aldrich). Following separation, a Bradford assay was conducted to determine the protein concentration in each fraction. Pyruvate levels were determined in cellular fractions and mitochondrial fractions by using the Pyruvate Assay Kit (#ab65342, Abcam) according to the manufacturer’s instructions. Briefly, after deproteinization using perchloric acid, the samples were neutralized in ice-cold 2 M KOH. 10 µL samples were incubated with reaction mix and kept on the plate at RT for 30 min in the dark. The absorbance was measured in a microplate reader (Tecan Group Ltd, Männedorf, Switzerland) at 570 nm, and the results were shown in three independent experiments.

Bacterial strains and growth conditions

The PA strain known as RifR human clinical isolate UCBPP-PA14 (also known as PA14) was used [70]. The bacteria were grown at 37 °C in lysogeny broth (LB) under shaking and aeration or on LB agar plates containing appropriate antibiotics. PA14 and isogenic mutant ΔmvfR [19] cultures were grown in LB from a single colony to an optical density of 600 nm (OD600) of 1.5, diluted 1:50,000,000 in fresh LB media, and grown overnight to an OD600 of 3.0.

Gentamicin protection assay

RAW 264.7 macrophage cells were plated on 6 well-cell culture-treated plates overnight in pyruvate-free DMEM. After 3 h of incubation with 10 µM UK5099, 20 µM ATP or 400 µM 2-AA, cells were infected with PA14 and isogenic mutant ΔmvfR at 5 MOI for 30 min at 37 °C in 5% CO2. Unbound bacteria were removed by washing once with cold DMEM medium. Afterward, the cells were incubated with 100 µg/ml of gentamicin for 30 min to eliminate residual extracellular bacteria. The cells were washed with DMEM, transferred to free medium without gentamicin and kept for 3 h at 37°C in 5% CO2. The infected cells were scraped after 3 h, centrifuged at 500 × g, and lysed in distilled water. The lysed cells were immediately diluted in PBS and plated on LB agar plates to assess bacterial presence. Bacterial colony-forming units (CFUs) were counted after incubating the plates overnight at 37°C.Untreated cells infected with PA14 were set as 100 % and the reduction of the bacterial load was expressed as %CFU.

Pharmacological inhibitors and ATP supplementations

For the OXPHOS inhibition assay, RAW 264.7 macrophage cells were treated with UK5099 (10 µM, Sigma-Aldrich, dissolved in DMSO 3 h prior to 1st or 2nd rounds of 2-AA exposure. RAW 264.7 macrophages supplemented with ATP (20 µM, Sigma-Aldrich, dissolved in PBS) received ATP at the time of 1st and 2nd 2-AA exposure.

Co-Immunoprecipitation assay

For all immunoprecipitation assays, protein A/G agarose Magnetic beads (Pierce) were used after washing in 1X IP buffer. Nuclear lysates from RAW 264.7 tolerized cells exposed to 2-AA for 24 h or non-exposed cells were diluted in 1X IP buffer, and 500 µg of protein was taken for each experiment. The lysates were precleared by using unbound 50 µl protein A/G magnetic beads for 2 h at room temperature on a rotator. Precleared lysates were either incubated with ERRα, PGC-1α or rabbit IgG antibody overnight at 4 °C. 100 µl protein A/G magnetic beads were used to pull down the antibody-protein complex and washed twice with IP buffer to remove unbound proteins. Magnetic beads were then eluted in 2X Laemmli SDS-PAGE loading buffer. Immunoprecipitated ERRα and PGC-1α was detected by Western blot analyses, using conformational specific Anti Rabbit antibody (TruBlot).

Immunoblotting analysis

Cells were plated at 6×105 cells per well in six-well plates. Cells were washed with PBS and subsequently lysed using RIPA lysis buffer containing 1 mM phenylmethylsulfonyl fluoride (PMSF). 20μg of proteins were separated by electrophoresis on any KD (Kilo Dalton) (Bio-Rad, Cat no. 4569033) SDS-polyacrylamide gel. Proteins were transferred to a 0.2 μm polyvinylidene fluoride membrane (PVDF, Millipore, Billerica, USA) using a Bio-Rad semi-dry instrument. After blocking with 5% BSA in TBS containing 0.1% Tween-20 for 1 h at room temperature, the membranes were incubated with a primary antibody ERR-a (Abcam, #ab76228), MPC1 (D2L91, Cell signaling), and anti-β-actin (cat no: sc-47778) (Santa Cruz Biotechnology) overnight at 4°C. Following washing, the membranes were incubated with an anti-rabbit secondary antibody, and the bands were detected by SuperSignal West Pico Chemiluminescent Substrate (Thermo Scientific) reaction, according to the manufacturer’s instructions. The blots were visualized in the ChemiDOC Imaging system (Bio-Rad Laboratories, Inc., Hercules, CA, USA). The bands were analyzed densitometrically using QuantityOne software (Bio-Rad).

Chromatin immunoprecipitation (ChIP) and ChIP-qPCR

For ChIP studies, RAW 264.7 macrophage cells exposed to 2-AA for 24 h were cross-linked in 1% (v/v) methanol-free formaldehyde for 10 min and then placed in 0.125 M glycine for 5 min at RT. Using the truChIP High Cell Chromatin Shearing kit (Covaris, United States), cells were prepared for sonication according to the Covaris protocol. Approximately 1 × 107 cells were plated in a 12 mm × 12 mm tube and subjected to shearing with the Covaris S220 sonicator for 8 min (140 peak power, 5 duty factor, 200 cycles/burst). The Magna ChIP A/G kit (Millipore, United States) was used for the subsequent immunoprecipitations according to the manufacturer’s protocol. Briefly, chromatin from approximately 106 cells was incubated overnight at 4°C with 2 μg of anti-ERRα or anti-PGC-1α (Abcam, USA) ChIP-grade antibody and 20 μL of A/G magnetic beads. The beads were washed serially (5 min each) with low-salt wash buffer, high-salt wash buffer, LiCl wash buffer, and TE buffer from the kit at 4°C. Chromatin was eluted with elution buffer containing Proteinase K at 62°C for 4 h, then incubated at 95°C for 10 min. DNA was isolated by column purification (QIAquick PCR purification kit).

Real-time ChIP-qPCR was performed with the Brilliant II SYBR green super mix (Agilent, USA). Forward (AGTGGTGACCTTGAACTTCCC) and reverse (CTGAAGACGACCTTCCCCTT) primers were chosen to amplify a genomic locus of MPC1 promoter, which had a putative ERR binding site (TNAAGGTCA) at 1582 bp upstream of the start site. Normalized values were calculated using the percent-input method relative to the IgG. The assay was performed three times.

RNA extraction and RT-qPCR

Total RNA from all the groups mentioned above was isolated from approximately 2 × 106 cells with the RNeasy minikit (Qiagen, USA), and cDNA was prepared with the iScript Reverse transcription kit (Bio-Rad, USA), as per the manufacturer’s instruction. Real-time PCR was conducted using the PowerUP SYBR Green Master mix (Applied Biosystem, USA) and primer sets for mouse ERR-α (forward: ACTACGGTGTGGCATCCTGTGA; reverse: GGTGATCTCACACTCATTGGAGG), PGC-1α (forward: GAATCAAGCCACTACAGACACCG; reverse: CATCCCTCTTGAGCCTTTCGTG), MPC-1 (forward: CTCCAGAGATTATCAGTGGGCG; reverse: GAGCTACTTCGTTTGTTACATGGC), TNF-α (forward: GGTGCCTATGTCTCAGCCTCTT; reverse: GCCATAGAACTGATGAGAGGGAG) and mouse 18S rRNA (forward: GTTCCGACCATAAACGATGCC; reverse: TGGTGGTGCCCTTCCGTCAAT). The transcript levels of all the genes were normalized to 18S rRNA with the ΔΔCT method. The relative expression was calculated by normalizing transcript levels to those of PA14-infected cells. The assay was conducted in triplicate; means and standard deviations were calculated for each group.

Animal infection and metabolites assessment experiments

The full-thickness thermal burn injury and infection (BI) model [17, 18, 71] was used to assess the effect of 2-AA on metabolic alterations in the spleens of six-week-old CD1 male mice and bacterial burden (Charles River Labs, USA). A full-thickness thermal burn injury involving 5-8% of the total body surface area was produced on the shaved mouse abdomen dermis, and an inoculum of ∼1 × 104 PA14 and ΔmvfR [19] cells in 100 μL of MgSO4 (10 mM) was injected intradermally into the burn eschar. For the groups that received 2-AA, mice were injected intraperitoneally (IP) with 100 μL of 2-AA (6.75 mg/kg). The entire procedure was done under the influence of anesthesia. One of the groups of mice infected with the PA14 isogenic mutant ΔmvfR [19] also received 100 μL of 2-AA in PBS (6.75 mg/kg) at the time of infection (ΔmvfR +2-AA) and served as an additional control. CFU counts from rectus abdominus muscle (underlying the burn-infected tissue) were assessed in groups of 4 mice each at 1, 5, and 10 d post-BI by plating diluted muscle homogenate on Pseudomonas isolation agar (Sigma-Aldrich) plates containing rifampicin (50 mg/L). Spleen samples from mice were collected for metabolite analyses from all mice groups at 1, 5, and 10 d post-BI to assess ATP (#ab83355 Abcam) and acetyl-CoA (#ab87546, Abcam) as described above. Tissues were homogenized, 100 mg of homogenate was centrifuged at 4°C for 15 min at 10,000 g, and supernatants were collected. The supernatant was mixed with 400 µL of 1 M perchloric acid. The deproteinized supernatant was neutralized by 3 M KHCO3. The ATP and acetyl-CoA assays were performed as described in the in vitro section.

Electron microscopy studies

RAW 264.7 macrophages exposed to 2-AA and corresponding controls were fixed with 2% glutaraldehyde in 0.1 M cacodylate buffer and postfixed in 1% OsO4 in 0.1 M cacodylate buffer for 1 h on ice. The cells were stained all at once with 2% uranyl acetate for 1 h on ice, after which they were dehydrated in a graded series of ethanol (50–100%) while remaining on ice. Ultrathin (70 nm) sections were cut using a Leica EMUC7 ultramicrotome and collected onto formvar-coated grids (EMS, Hatfield, PA). Sections were contrast-stained using 2.0% aqueous uranyl acetate. Grids were examined at 80 kV in a JEOL 1011 transmission electron microscope (Peabody, MA) equipped with an AMT digital camera and proprietary image capture software (Advanced Microscopy Techniques, Danvers, MA).